---
title: "Homework 1"
subtitle: "BSTA 512"
author: "Your name here - update this!!!!"
description: "Due: Thursday January 18, 2023"
date-modified: "12/22/2023"
date: "1/18/2024"
categories: [""]
format: 
  html: 
    link-external-newwindow: true
    toc: true
  pdf: default 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 2, fig.width = 4)

library(tidyverse)
library(rstatix)
library(broom)
library(gt)
library(janitor)
```

## Directions

-   Please upload your homework to Sakai. **Upload both your .Rmd code file and the knitted .html file**.

-   For each question, make sure to include all code and resulting output in the html file to support your answers.

-   Show the work of your calculations using R code within a code chunk. Make sure that both your code and output are visible in the knitted html file.

-   Write all answers in complete sentences as if communicating the results to a collaborator.

::: callout-tip
It is a good idea to try rendering your document from time to time as you go along! Note that rendering automatically saves your qmd file and rendering frequently helps you catch your errors more quickly.
:::

## Questions

The following questions were adapted from this textbook.

-   Download the datasets from Sakai - see Week 2

### Question 1

-   Load the dataset using the `readxl` package.
    -   This `readxl` package was installed as a part of the tidyverse, however it does not get loaded when you load the tidyverse package and thus you need to do that separately.
    -   Use the command `read_excel()`, as shown below

```{r}
library(readxl)
# you might need to update the location of the data file
# you can choose whatever name you like for the tibble when loading it into R's workspace 
ch05q01 <- read_excel("data/CH05Q01.xls")
```

### (a)

<font color="green">Additional instructions: Create the scatterplots in R using ggplot, in addition to describing the relationships in the book's instructions.</font>

### (b)

<font color="green">Note: This is asking for the regression models BEFORE you find the values of the coefficients.</font>

### (c)

<font color="green">Note: Now get the regression coefficients using R and plug them into the regression models from (b). You can get the coefficients from the R output - you don't have to use the formulas.</font>

### (d)

<font color="green">Note: Instead of sketching the lines, create them using ggplot.</font>

### (e)

<font color="green">Note: You can get the CI's from the R output - you don't have to use the formulas.</font>

### (f)

<font color="green">Note: Instead of sketching, make the figure using ggplot. Get the asked for CI using R instead of estimating it based on the sketch.</font>

## 5.4 (a - g)

```{r}
ch05q04 <- read_excel("data/CH05Q04.xls")
```

### (a)

<font color="green">Additional instructions: Run the model to get the regression coefficients and create a scatterplot with the regression line using ggplot.</font>

### (b)

### (c)

<font color="green">Additional instructions: Calculate the CI using the formula. The problem gives the values for $S_{Y|X}$ and $S_X$. Verify these values using R.</font>

### (d)

### (e)

<font color="green">Additional instructions: Create a new dataset without the outlier. Verify the regression coefficients using R and create a new scatterplot with regression line for the dataset without the outlier. Then "decide whether this outlier has any effect on your estimate of the IQ–DI relationship."</font>

### (f)

<font color="green">Additional instructions: Calculate the test statistic using the formula and find the p-value using the test statistic value. The problem gives the values for $S_{Y|X}$ and $S_X$. Verify these values using R. Check your work by comparing your answers to the linear model output. </font>

### (g)
