---
title: "Homework 1"
subtitle: "BSTA 512/612"
author: "Your name here - update this!!!!"
description: "Due: Thursday January 25, 2023"
date-modified: "1/16/2024"
date: "1/25/2024"
format: 
  html: 
    link-external-newwindow: true
    toc: true
    embed-resources: true
    self-contained: true
  pdf: default 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 2, fig.width = 4)

library(tidyverse)
library(rstatix)
library(broom)
library(gt)
library(janitor)
library(here)

```

::: callout-important
**This page is no longer under construction. You may start this homework! (1/16/24)**
:::

## Directions

-   [Download the `.qmd` file here.](https://github.com/nwakim/W2024_BSTA_512/blob/main/homework/HW1.qmd)

-   You will need to download the datasets. Use [this link to download](https://ohsuitg-my.sharepoint.com/:f:/r/personal/wakim_ohsu_edu/Documents/Teaching/Classes/W2024_BSTA_512_612/Student_files/Data/Homework/HW1?csf=1&web=1&e=fPDqYu) the HW1 datasets needed in this assignment. If you do not want to make changes to the paths set in this document, then make sure the files are stored in a folder named "data" that is housed in the same location as your HW1 `.qmd` file.

-   Please upload your homework to Sakai. **Upload both your `.qmd` code file and the rendered `.html` file**

-   For each question, make sure to include all code and resulting output in the html file to support your answers

-   Show the work of your calculations using R code within a code chunk. Make sure that both your code and output are visible in the rendered html file. This is the default setting.

-   Write all answers in complete sentences as if communicating the results to a collaborator.

::: callout-tip
It is a good idea to try rendering your document from time to time as you go along! Note that rendering automatically saves your `.qmd` file and rendering frequently helps you catch your errors more quickly.
:::

## Questions

The following questions were adapted from [this textbook](https://catalog.lib.uchicago.edu/vufind/Record/7546272).

### Question 1

Please use R code to determine the following answers. *(adapted from problem 3.3 in Applied Regression Analysis and Other Multivariable Methods)*

::: {.callout-note appearance="simple"}
Type `?pnorm` in the console to get some information on a potentially helpful function.
:::

#### Part a

From a normal distribution with mean 4 and standard deviation 6, what is $P(X>2)$?

#### Part b

From a normal distribution with mean 4 and standard deviation 6, for what value (in place of ??) would $P(X>??) = 0.1$?

### Question 2

Suppose that the height ($H$) of assigned-male-at-birth (AMAB) patients registered at a clinic has the normal distribution with mean 70 inches and variance 4. *(adapted from problem 3.11 in Applied Regression Analysis and Other Multivariable Methods)*

#### Part a

For a random sample of patients of size $n = 25$, the expression $P(\bar{H} < 65)$, in which $\bar{H}$ denotes the sample mean height, is equivalent to saying $P(Z < ?)$

::: {.callout-note appearance="simple"}
$Z$ is a standard normal random variable.
:::

#### Part b

Using the `pnorm` function, show that the probability expressions in Part a are equal.

#### Part c

Find an interval $(a, b)$ such that $P(a< \bar{H} <b) = 0.80$ for the same random sample in Part a.

### Question 3

Test the null hypothesis that the true population average height is the same for two independent groups from one hospital versus the alternative hypothesis that these two population averages are different, using the following data:

-   Group 1: \[69.25, 72.80, 68.73, 72.01, 70.36, 71.49, 72.73\]

-   Group 2: \[67.54, 68.51, 71.84, 70.59, 71.52, 71.50\]

You may assume that the populations from which the data come are each normally distributed, with equal population variances. What conclusion should be drawn, with $\alpha = 0.05$?

::: {.callout-important appearance="simple"}
**Please attempt this problem using R.** Take a look at the information for the `t.test` function. You will need to set `x`, `y`, `alternative`, and `var.equal=T`. You can use the below groups coded in R.
:::

```{r}
grp1 = c(69.25, 72.80, 68.73, 72.01, 70.36, 71.49, 72.73)
grp2 = c(67.54, 68.51, 71.84, 70.59, 71.52, 71.50)
```

### Question 4

The choice of an alternative hypothesis ($H_A$ or $H_1$) should depend primarily on (choose all that apply). Explain your reasoning.

a.  the data obtained from the study.

b.  what the investigator is interested in determining.

c.  the critical region.

d.  the significance level.

e.  the power of the test.

### Question 5

[Visit this site on dplyr.](https://dplyr.tidyverse.org/reference/)

For one of the functions that we have not discussed in class, please use it on the `dds.discr` dataset. Please write in words what you would like to perform, then write the code.

Note: the `dds.discr` dataset is an `.Rda` file. Instead of using `read_csv()` or `read_excel()`, you can use `load()`.

### Question 6

The accompanying data `CH05Q01` gives the dry weights (Y) of 11 chick embryos ranging in age from 6 to 16 days (X ). Also given in the data are the values of the common logarithms of the weights (Z).

-   Load the dataset using the `readxl` package.
    -   This `readxl` package was installed as a part of the tidyverse, however it does not get loaded when you load the tidyverse package and thus you need to do that separately.
    -   Use the command `read_excel()`, as shown below

```{r}
library(readxl)
# you might need to update the location of the data file
# you can choose whatever name you like for the tibble when loading it into R's workspace 
ch05q01 <- read_excel("./data/CH05Q01.xls")
```

#### Part a

Create two scatterplots in R using ggplot:

1.  Between age (X) and dry weight (Y)

2.  Between age (X) and log10 dry weight (Z)

Observe the following two scatter diagrams. Describe the relationships between age (X) and dry weight (Y) and between age and log10 dry weight (Z).

#### Part b

State the population simple linear regression models for these two regressions: Y regressed on X and Z regressed on X.

::: {.callout-note appearance="simple"}
This is asking for the regression models BEFORE you find the values of the coefficients.
:::

#### Part c

Determine the least-squares estimates of each of the regression lines in part (b).

::: {.callout-note appearance="simple"}
Now get the regression coefficients using R and plug them into the regression models from (b). You can get the coefficients from the R output!
:::

#### Part d

Using `ggplot`, create the respective best-fit lines on your plots. Which of the two regression lines has the better fit? Based on your answers to parts (a)–(c), is it more appropriate to run a linear regression of Y on X or of Z on X? Explain.

#### Part e (to be covered on Monday 1/22)

For the regression that you chose as being more appropriate in part (d), find 95% confidence intervals for the true slope and intercept. Interpret each interval ~~with regard to the null hypothesis that the true value is 0~~.

::: {.callout-note appearance="simple"}
You can get the CI's from the R output
:::

::: callout-important
## Update 1/21/2024

I removed some of the instructions because they were a little confusing. No need to go through the formal process of a hypothesis test.
:::

#### Part f (to be covered on Monday 1/22)

For the regression that you chose as being more appropriate in part (d), add 95% confidence ~~and prediction~~ bands. Using your sketch, find and interpret an approximate 95% confidence interval for the mean response of an 8-day-old chick.

::: callout-important
## Update 1/21/2024

I removed prediction bands because we did not cover them!
:::

::: callout-note
## Update 1/22/2024

You do NOT need to do this problem (Question 6, part f). We did NOT get to it on Monday, so I don't expect you to do it for the homework.
:::

## Question 7

Go to the [R Documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm) of the `lm()` function. Please answer the following questions about the function arguments and output.

### Part a

If you wanted to perform weighted least squares in R, what argument would you need to change?

### Part b

What is the default method in `lm()`? Can it be used to solve OLS? (Hint: check out the Wiki page for the method)

### Part c

In linear regression, a singular fit means that the estimated standard deviation of your residuals is very close to zero. In the `lm()` function, will you get an error if you have a singular fit?

## Question 8

**Quick True/False question:** Is ordinary least squares the only way to find the best fit line for linear regression?
