---
title: "Homework 1"
subtitle: "BSTA 512/612"
author: "Your name here - update this!!!!"
description: "Due: Thursday January 18, 2023"
date-modified: "12/22/2023"
date: "1/18/2024"
categories: [""]
format: 
  html: 
    link-external-newwindow: true
    toc: true
    embed-resources: true
  pdf: default 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 2, fig.width = 4)

library(tidyverse)
library(rstatix)
library(broom)
library(gt)
library(janitor)
library(here)

```

::: callout-important
**THIS PAGE IS UNDER CONSTRUCTION!!**
:::

## Directions

-   Please upload your homework to Sakai. **Upload both your .Rmd code file and the knitted .html file**.

-   For each question, make sure to include all code and resulting output in the html file to support your answers.

-   Show the work of your calculations using R code within a code chunk. Make sure that both your code and output are visible in the knitted html file.

-   Write all answers in complete sentences as if communicating the results to a collaborator.

::: callout-tip
It is a good idea to try rendering your document from time to time as you go along! Note that rendering automatically saves your qmd file and rendering frequently helps you catch your errors more quickly.
:::

## Questions

The following questions were adapted from this textbook.

-   Download the datasets from Sakai - see Week 2

## Question 1

Visit this site on dplyr. <https://dplyr.tidyverse.org/reference/index.html#groups>

For one of the functions that we have not discussed in class, please use it on the `dds.discr` dataset.

### Question 1

Please use R code to determine the following answers. *(adapted from problem 3.3)*

::: {.callout-note appearance="simple"}
Type `?pnorm` in the console to get some information on a potentially helpful function.
:::

#### Part a

From a normal distribution with mean 4 and standard deviation 6, what is $P(X>2)$?

#### Part b

From a normal distribution with mean 4 and standard deviation 6, for what value (in place of ??) would $P(X>??) = 0.1$?

### Question 2

Suppose that the height ($H$) of assigned-male-at-birth (AMAB) patients registered at a clinic has the normal distribution with mean 70 inches and variance 4. *(adapted from problem 3.11)*

#### Part a

For a random sample of patients of size $n = 25$, the expression $P(\bar{H} < 65)$, in which $\bar{H}$ denotes the sample mean height, is equivalent to saying $P(Z < ?)$

::: {.callout-note appearance="simple"}
$Z$ is a standard normal random variable.
:::

#### Part b

Using the `pnorm` function, show that the probability expressions in Part a are equal.

#### Part c

Find an interval $(a, b)$ such that $P(a< \bar{H} <b) = 0.80$ for the same random sample in Part a.

### Question 3

Test the null hypothesis that the true population average height is the same for two independent groups from one hospital versus the alternative hypothesis that these two population averages are different, using the following data:

-   Group 1: \[69.25, 72.80, 68.73, 72.01, 70.36, 71.49, 72.73\]

-   Group 2: \[67.54, 68.51, 71.84, 70.59, 71.52, 71.50\]

You may assume that the populations from which the data come are each normally distributed, with equal population variances. What conclusion should be drawn, with $\alpha = 0.05$?

::: {.callout-important appearance="simple"}
**Please attempt this problem using R.** Take a look at the information for the `t.test` function. You will need to set `x`, `y`, `alternative`, and `var.equal=T`. You can use the below groups coded in R.
:::

```{r}
grp1 = c(69.25, 72.80, 68.73, 72.01, 70.36, 71.49, 72.73)
grp2 = c(67.54, 68.51, 71.84, 70.59, 71.52, 71.50)
```

### Question 4

The choice of an alternative hypothesis ($H_A$ or $H_1$) should depend primarily on (choose all that apply)

a.  the data obtained from the study.

b.  what the investigator is interested in determining.

c.  the critical region.

d.  the significance level.

e.  the power of the test.

### Question 5

The accompanying table gives the dry weights (Y ) of 11 chick embryos ranging in age from 6 to 16 days (X ). Also given in the table are the values of the common logarithms of the weights (Z).

-   Load the dataset using the `readxl` package.
    -   This `readxl` package was installed as a part of the tidyverse, however it does not get loaded when you load the tidyverse package and thus you need to do that separately.
    -   Use the command `read_excel()`, as shown below

```{r}
library(readxl)
# you might need to update the location of the data file
# you can choose whatever name you like for the tibble when loading it into R's workspace 
ch05q01 <- read_excel("./data/CH05Q01.xls")
```

### Part a

Create the scatterplots in R using ggplot the above dataset. Observe the following two scatter diagrams. Describe the relationships between age (X) and dry weight (Y) and between age and log10 dry weight (Z).

### Part b

State the simple linear regression models for these two regressions: Y regressed on X and Z regressed on X.

::: {.callout-note appearance="simple"}
This is asking for the regression models BEFORE you find the values of the coefficients.
:::

### Part c

Determine the least-squares estimates of each of the regression lines in part (b).

::: {.callout-note appearance="simple"}
Now get the regression coefficients using R and plug them into the regression models from (b). You can get the coefficients from the R output - you don't have to use the formulas.
:::

### Part d

Create a line on your plots. Which of the two regression lines has the better fit? Based on your answers to parts (a)–(c), is it more appropriate to run a linear regression of Y on X or of Z on X? Explain.

### Part e

For the regression that you chose as being more appropriate in part (d), find 95% confidence intervals for the true slope and intercept. Interpret each interval with regard to the null hypothesis that the true value is 0.

::: {.callout-note appearance="simple"}
You can get the CI's from the R output - you don't have to use the formulas.
:::

### Part f

For the regression that you chose as being more appropriate in part (d), add 95% confidence and prediction bands. Using your sketch, find and interpret an approximate 95% confidence interval for the mean response of an 8-day-old chick.

## 5.4 (a - g)

```{r}
ch05q04 <- read_excel("./data/CH05Q04.xls")
```

### (a)

<font color="green">Additional instructions: Run the model to get the regression coefficients and create a scatterplot with the regression line using ggplot.</font>

### (b)

### (c)

<font color="green">Additional instructions: Calculate the CI using the formula. The problem gives the values for $S_{Y|X}$ and $S_X$. Verify these values using R.</font>

### (d)

### (e)

<font color="green">Additional instructions: Create a new dataset without the outlier. Verify the regression coefficients using R and create a new scatterplot with regression line for the dataset without the outlier. Then "decide whether this outlier has any effect on your estimate of the IQ–DI relationship."</font>

### (f)

<font color="green">Additional instructions: Calculate the test statistic using the formula and find the p-value using the test statistic value. The problem gives the values for $S_{Y|X}$ and $S_X$. Verify these values using R. Check your work by comparing your answers to the linear model output. </font>

### (g)
