---
title: "Homework 3"
subtitle: "BSTA 512/612"
author: "Your name here!!!"
description: "Due: Thursday February 15, 2024 at 11pm"
date-modified: "2/15/2024"
date: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
  pdf: default 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 2, fig.width = 4)

library(tidyverse)
library(rstatix)
library(broom)
library(gt)
library(janitor)
library(readxl)
```

::: callout-important
**THIS PAGE IS UNDER CONSTRUCTION!!**
:::

## Directions

-   [Download the `.qmd` file here.](https://github.com/nwakim/W2024_BSTA_512/blob/main/homework/HW3.qmd)

-   You will need to download the datasets. Use [this link to download](https://ohsuitg-my.sharepoint.com/:f:/r/personal/wakim_ohsu_edu/Documents/Teaching/Classes/W2024_BSTA_512_612/Student_files/Data/Homework?csf=1&web=1&e=QEsqb9) the homework datasets needed in this assignment. If you do not want to make changes to the paths set in this document, then make sure the files are stored in a folder named "data" that is housed in the same location as this homework `.qmd` file.

-   Please upload your homework to [Sakai](https://sakai.ohsu.edu/portal/site/BSTA-512-1-AB-W24/tool/961ec5ef-d7f9-4518-be16-25f7af3d6be4). **Upload both your `.qmd` code file and the rendered `.html` file**

    -   **Please rename you homework as `Lastname_Firstinitial_HW0.qmd`. This will help organize the homeworks when the TAs grade them.**

    -   **Please also add the following line under `subtitle: "BSTA 512/612"`: `author: First-name Last-name` with your first and last name so it is attached to the viewable document.**

-   For each question, make sure to include all code and resulting output in the html file to support your answers.

-   Show the work of your calculations using R code within a code chunk. Make sure that both your code and output are visible in the rendered html file. This is the default setting.

-   If you are computing something by hand, you may take a picture of your work and insert the image in this file. You may also use LaTeX to write it inline.

-   Write all answers in complete sentences as if communicating the results to a collaborator. This means including a sentence summarizing results in the context of the research study.

::: callout-tip
It is a good idea to try rendering your document from time to time as you go along! Note that rendering automatically saves your qmd file and rendering frequently helps you catch your errors more quickly.
:::

## Questions

### Question 1

A high respiratory rate is a potential diagnostic indicator of respiratory infection in children. To judge whether a respiratory rate is “high” however, a physician must have a clear picture of the distribution of normal rates. To this end, Italian researchers measured the respiratory rates of 618 children between the ages of 15 days and 3 years.

The data and problem framing came from the `Sleuth3` package. Please make sure to run the following code to load the data. You can directly access the dataset `ex0824` from the package. I have included a new assignment of the data to `q1_data` if you would like to use that.

```{r}
#| message: false

if(!require(Sleuth3)) { install.packages("Sleuth3"); library(Sleuth3) }
q1_data = ex0824
```

#### Part a

Create a scatterplot of the dependent and independent variables with both the best-fit line and a smoothed curve through the points. Describe the relationship between the dependent and independent variables, and also comment on whether you think it is reasonable to use a linear regression to model the relationship. Are there any outliers in the data? If so, describe the points and why you think they are outliers.

## (2)

Write out the regression equation for the simple linear regression model.

## (3)

Assess the normality of the model's (ordinary) residuals by creating a histogram, density plot, and boxplot of the residuals to visually inspect the distribution of the residuals, and describe any deviations from normality.

## (4)

Assess the normality of the model's (ordinary) residuals by creating a normal probability plot of the residuals. Compare the normality probability plot to 8 such plots simulated from normal data, and discuss why or why not the residuals could have come from a normal distribution.

## (5)

Test the normality of the model's (ordinary) residuals and comment on whether the test's conclusion is consistent with your visual inspection or not. Make sure to include the hypotheses and a conclusion to the test based on the p-value.

## (6)

Create a residual plot using ggplot and the standardized residuals and discuss what this shows us in terms of whether the model assumptions have been met or not.

## (7)

Determine whether there are any observations with high leverage. If there are observations with high leverage, identify their coordinates and describe how they relate to the other observations. Why would these points have high leverage compared to the other observations? Do you think removing the points would change the linear model much? (you do not need to remove the points and rerun the model, just comment on whether you think they are influential)

## (8)

Determine whether there are any observations with high Cook's distance. If there are observations with high Cook's distance, identify their coordinates and describe how they relate to the other observations. Why would these points have high Cook's distance compared to the other observations? Do you think removing the points would change the linear model much? (you do not need to remove the points and rerun the model, just comment on whether you think they are influential)

## (9)

Create histograms and density plots of the dependent and independent variables and describe their distribution shapes.

## (10)

Use Tukey's ladder of transformations to choose two possible transformations for the dependent variable. Explain why you chose them. Note: questions below will ask about model fit with the transformations. For now, just explain why you chose the ones that you did.

## (11)

Use Tukey's ladder of transformations to choose two possible transformations for the independent variable. Explain why you chose them. Note: questions below will ask about model fit with the transformations. For now, just explain why you chose the ones that you did.

## (12)

Add the 4 transformations you chose above (2 for the dependent variable and 2 for the independent variable) to the dataset.

## (13)

Create scatterplots using the transformed variables and discuss whether any of the transformations improve the model fit and why (or why not). Include plots with just the x or y variables transformed, and at least one plot with both the x and y variables transformed.

## (14)

Run the various transformed models and save the output to use for the diagnostic questions below.

## (15)

Compare the normal QQ plots of the different models and discuss whether any of the transformations improve the model fit and why (or why not).

## (16)

Compare the residual plots of the different models and discuss whether any of the transformations improve the model fit and why (or why not).

## (17)

Compare the leverage & Cook’s distance of the different models and discuss whether any of the transformations improve the model fit and why (or why not).

## (18)

Compare the $R^2$ values and F-test p-values of the different models and discuss whether any of the transformations improve the model fit and why (or why not).

## (19)

Which of the models would you recommend using for analyses? Discuss why you chose the model and why you did not choose the other models.

# Problem 9.7 (a, b, e, f) from book

*Note: (c) is in recommended problems below.*

> Use the results from Problem 4 in Chapter 8, as well as the computer output given here, to answer the following questions about the data from that problem.

Note: Create all output you need using R instead of relying on the output given in the text.

Problem 4 from Chapter 8:

> A sociologist investigating the recent increase in the incidence of homicide throughout the United States studied the extent to which the homicide rate per 100,000 population ($Y$) is associated with the city’s population size ($X_1$), the percentage of families with yearly income less than \$5,000 ($X_2$), and the rate of unemployment ($X_3$). Data are provided in the following table for a hypothetical sample of 20 cities.

## (a)

> Conduct the overall regression F test for the model where $Y$ is regressed on $X_1, X_2$, and $X_3$. Use alpha = 0.05. Interpret your result.

## (b)

> Provide variables-added-in-order tests for the order $X_2, X_1$, and $X_3$.

This means that there are 3 tests: (1) test model with just $X_2$, (2) test adding $X_1$ to the model given that $X_2$ is already in the model, and (3) test adding $X_3$ to the model given that $X_2, X_1$ are already in the model. See the subsections below to divide up the work.

### Test model with just $X_2$

### Adding $X_1$ to the model given that $X_2$ is already in the model

### Adding $X_3$ to the model given that $X_2, X_1$ are already in the model

## (e)

> Provide variables-added-last tests for $X_1, X_2$, and $X_3$.

This means that there are 3 tests: (1) test adding $X_1$ to the model given that $X_2, X_3$ are already in the model, (2) test adding $X_2$ to the model given that $X_1, X_3$ are already in the model, and (3) test adding $X_3$ to the model given that $X_1, X_2$ are already in the model. See the subsections below to divide up the work.

-   For adding $X_2$ and $X_3$ last to the model, you do not need to calculate the test statistic using the formula or the p-value directly using its probability distribution. You can instead run the appropriate tests in R.

### Adding $X_1$ to the model given that $X_2, X_3$ are already in the model

### Adding $X_2$ to the model given that $X_1, X_3$ are already in the model

### Adding $X_3$ to the model given that $X_1, X_2$ are already in the model

## (f)

> Provide the variables-added-last test for $X_4 = X_2 X_3$ given that $X_2$ and $X_3$ are already in the model. Does $X_4$ significantly improve the prediction of $Y$ given that $X_2$ and $X_3$ are already in the model?
