---
title: "Homework 5"
subtitle: "BSTA 512/612"
author: "Your name here!!!"
description: "Due: Thursday March 7, 2024 at 11pm"
date-modified: "today"
date: "3/7/2024"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    embed-resources: true
---

::: callout-important
This assignment is under construction !!! (2/29/2024)
:::

```{=html}
<style>div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 6, fig.align = "center")

library(tidyverse)
library(rstatix)
library(broom)
library(gt)
library(janitor)
library(readxl)
library(gridExtra)
library(ggfortify)  # autoplot(model)
library(gtsummary)
library(describedata) # gladder()
library(haven)
library(here)
```


## Directions

-   [Download the `.qmd` file here.](https://github.com/nwakim/W2024_BSTA_512/blob/main/homework/HW5.qmd)

-   You will need to download the datasets. Use [this link to download](https://ohsuitg-my.sharepoint.com/:f:/r/personal/wakim_ohsu_edu/Documents/Teaching/Classes/W2024_BSTA_512_612/Student_files/Data/Homework?csf=1&web=1&e=QEsqb9) the homework datasets needed in this assignment. If you do not want to make changes to the paths set in this document, then make sure the files are stored in a folder named "data" that is housed in the same location as this homework `.qmd` file.

-   Please upload your homework to [Sakai](https://sakai.ohsu.edu/portal/site/BSTA-512-1-AB-W24/tool/961ec5ef-d7f9-4518-be16-25f7af3d6be4). **Upload both your `.qmd` code file and the rendered `.html` file**

    -   **Please rename you homework as `Lastname_Firstinitial_HW5.qmd`. This will help organize the homeworks when the TAs grade them.**

    -   **Please also add the following line under `subtitle: "BSTA 512/612"`: `author: First-name Last-name` with your first and last name so it is attached to the viewable document.**

-   For each question, make sure to include all code and resulting output in the html file to support your answers.

-   Show the work of your calculations using R code within a code chunk. Make sure that both your code and output are visible in the rendered html file. This is the default setting.

-   If you are computing something by hand, you may take a picture of your work and insert the image in this file. You may also use LaTeX to write it inline.

-   Write all answers in complete sentences as if communicating the results to a collaborator. This means including a sentence summarizing results in the context of the research study.

::: callout-tip
It is a good idea to try rendering your document from time to time as you go along! Note that rendering automatically saves your qmd file and rendering frequently helps you catch your errors more quickly.
:::

# Questions

The purpose of these problems is to integrate what we have learned so far into the full process that we might follow to analyze data. This will help you see how what we've learned will connect as a single work flow. Some of the things we have learned that will be covered:

-   Choosing what to test
-   Interpretations of coefficients
-   F-test conclusions
-   Interactions

## Question 1

We are going to revisit the Palmer Penguins dataset from Homework 4. Choosing what to test, interpretations of coefficients, F-test conclusions, and interactions

For this problem we will be using the `penguins` dataset from the `palmerpenguins` R package. We will look at the association between flipper length of penguins (measured in mm) and specific species of penguins.

Description from help file:

> Includes measurements for penguin species, island in Palmer Archipelago, size (flipper length, body mass, bill dimensions), and sex.

[More info about the data.](https://allisonhorst.github.io/palmerpenguins/)

```{r}
# first install the palmerpenguins package
# install.packages("palmerpenguins")
library(palmerpenguins)
data(penguins)

# run the command below to learn more about the variables in the penguins dataset
# ?penguins
```

### Part a

Can use body mass (g) as the explanatory variable (variable of interest).

```{r}

```





