---
title: "Simple Linear Regression (SLR)"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#213c96"
date: "01/12/2023"
categories: ["Week 1"]
format: 
  revealjs:
    theme: [default, simple_NW.scss]
    toc: true
    toc-depth: 1
    toc-title: Class Overview
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Intro
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(janitor)
library(knitr)
library(broom)
library(rstatix)
library(gt)
```

## Let's start with an example

::: columns
::: {.column width="55%"}
```{r}
#| fig-height: 8
#| fig-width: 11

gapm <- read_csv("data/lifeexp_femlit_2011.csv")
ggplot(gapm, aes(x = female_literacy_rate_2011,
                 y = life_expectancy_years_2011)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, size = 3, colour="#F14124") +
  labs(x = "Female literacy rate (%)", 
       y = "Life expectancy (years)",
       title = "Relationship between life expectancy and \n the female literacy rate in 2011") +
    theme(axis.title = element_text(
        size = 30), 
        axis.text = element_text(
        size = 25), 
        title = element_text(
        size = 30))

```
:::

::: {.column width="45%"}
Average life expectancy vs. female literacy rate

-   Each point on the plot is for a different country

-   $x$ = country's adult female literacy rate

-   $y$ = country's average life expectancy (years)

-   Data are from Gapminder (2011)

:::
:::

::: heq
$$\begin{aligned}
\hat{\text{life expectancy}} & =  50.9 + 0.232\cdot\text{female literacy rate}
\end{aligned}$$
:::

## Questions we can ask with this model

## Dependent vs. Independent Variables

## Association vs. prediction

## Study Design

## Simple Linear Regression Model

The (population) regression line is denoted by:

::: heq
$$\begin{aligned}
Y & =  \beta_0 + \beta_1X + \epsilon \nonumber
\end{aligned}$$
:::

-   $\beta_0$ and $\beta_1$ are **unknown** population parameters

-   $\epsilon$ (epsilon) is the error about the line

    -   It is assumed to be a random variable with a...

    -   Normal distribution with mean 0 and constant variance $\sigma^2$

-   Our goal is to estimate $\beta_0$, $\beta_1$, and $\sigma^2$

    -   The point estimates based on a sample are denoted by
        $\hat{\beta_0}$, $\hat{\beta_1}$, and $\hat{\sigma^2}$

## "Linear" Models

## Model Components

## Interpretations

## Parameter estimation: best fit line

## Least squares model assumptions

## Estimate of variance??
