---
title: "Lesson 11: Interactions Continued"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#213c96"
date: "02/26/2024"
categories: ["Week 8"]
format: 
  revealjs:
    theme: [default, simple_NW.scss]
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Interactions 2
    html-math-method: mathjax
    highlight-style: ayu
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(knitr)
library(broom)
library(rstatix)
library(gt)
library(readxl)
#----------
# new packages
# install.packages("describedata")
library(describedata) # gladder()
library(gridExtra)   # grid.arrange()
library(ggfortify)  # autoplot(model)
# New Day 6
library(gtsummary)

# New Day 7
library(plotly) # for plot_ly() command
library(GGally) # for ggpairs() command 
library(ggiraphExtra)   # for ggPredict() command

# Load the data - update code if the csv file is not in the same location on your computer
# If you need to download the file, please go to ur shared folder under Data > Slides
gapm <- read_excel("data/Gapminder_vars_2011.xlsx", 
                   na = "NA")  # important!!!! 


gapm_sub <- gapm %>% # called it gapm2_sub3 to be consistent with Day 7 notes
  drop_na(LifeExpectancyYrs, FemaleLiteracyRate, four_regions, FoodSupplykcPPD) %>%
  mutate(four_regions = factor(four_regions, 
                               levels = c("africa", "americas", 
                                          "asia", "europe"), 
                               labels = c("Africa", "Americas", 
                                          "Asia", "Europe"))) %>%
  rename(income_levels = `World bank, 4 income groups 2017`) %>%
  mutate(income_levels1 = factor(income_levels, 
                                levels = c("Low income", 
                                           "Lower middle income", 
                                           "Upper middle income", 
                                           "High income")), 
         income_levels2 = relevel(factor(income_levels, 
                                levels = c("Low income", 
                                           "Lower middle income", 
                                           "Upper middle income", 
                                           "High income"), 
                                labels = c("Lower income", "Lower income", 
                                            "Higher income", "Higher income")), 
                                ref = "Lower income"))

```

# Learning Objective

::: lob
5.  Interpret the interaction component of a model with **two continuous covariates**, and how the main variable's effect changes.
:::

6.  When there are only two covariates in the model, test whether one is a confounder or effect modifier.

## Do we think food supply is an effect modifier for female literacy rate?

::: columns
::: {.column width="40%"}
-   We can start by visualizing the relationship between life expectancy and female literacy rate *by food supply*

-   Questions of interest: Does the effect of female literacy rate on life expectancy differ depending on food supply?

    -   This is the same as: Is food supply is an effect modifier for female literacy rate? Is food supply an effect modifier of the association between life expectancy and female literacy rate?

-   Let's run an interaction model to see!
:::

::: {.column width="60%"}
```{r fig.height=4, fig.width=6, warning=F, fig.align='center'}
#| echo: false
m_int_fs = lm(LifeExpectancyYrs ~ FemaleLiteracyRate*FoodSupplykcPPD, data = gapm_sub)
(mr1_2d = ggPredict(m_int_fs, interactive = T) )
```
:::
:::

## Model with interaction between *two continuous variables*

Model we are fitting:

$$ LE = \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \beta_3 FLR^c \cdot FS^c + \epsilon$$

-   $LE$ as life expectancy
-   $FLR^c$ as the **centered** around the mean female literacy rate (continuous variable)
-   $FS^c$ as the **centered** around the mean food supply (continuous variable)

```{r}
#| code-fold: true
#| code-summary: "Code to center FLR and FS"
#| echo: true
gapm_sub = gapm_sub %>% 
  mutate(FLR_c = FemaleLiteracyRate - mean(FemaleLiteracyRate), 
         FS_c = FoodSupplykcPPD - mean(FoodSupplykcPPD))
mean_FS = mean(gapm_sub$FoodSupplykcPPD) %>% round(digits = 0)
mean_FLR = mean(gapm_sub$FemaleLiteracyRate) %>% round(digits = 2)
```

In R:

```{r}
#| echo: true

m_int_fs = lm(LifeExpectancyYrs ~ FLR_c + FS_c + FLR_c*FS_c, data = gapm_sub)
```

OR

```{r}
#| echo: true

m_int_fs = lm(LifeExpectancyYrs ~ FLR_c*FS_c, data = gapm_sub)
```

## Displaying the regression table and writing fitted regression equation

```{r}
#| echo: true
tidy_m_fs = tidy(m_int_fs, conf.int=T) 
tidy_m_fs %>% gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 5)
```

$$ \begin{aligned}
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \widehat\beta_2 FS^c + \widehat\beta_3 FLR^c \cdot FS^c \\
\widehat{LE} = & `r round(tidy_m_fs$estimate[1],2)` + `r round(tidy_m_fs$estimate[2],2)` \cdot FLR^c + `r round(tidy_m_fs$estimate[3],2)` \cdot FS^c - 0.00001 \cdot FLR^c \cdot FS^c
\end{aligned}$$

## Comparing fitted regression lines for various food supply values

$$ \begin{aligned}
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \widehat\beta_2 FS^c + \widehat\beta_3 FLR^c \cdot FS^c \\
\widehat{LE} = & `r round(tidy_m_fs$estimate[1],2)` + `r round(tidy_m_fs$estimate[2],2)` \cdot FLR^c + `r round(tidy_m_fs$estimate[3],2)` \cdot FS^c - 0.00001 \cdot FLR^c \cdot FS^c
\end{aligned}$$

To identify different lines, we need to pick example values of Food Supply:

::: columns
::: {.column width="33%"}
::: FS1
::: FS1-title
Food Supply of `r mean_FS -1000` kcal PPD
:::

::: FS1-cont
$$\begin{aligned} 
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \\ & \widehat\beta_2 \cdot (-1000) + \\ & \widehat\beta_3 FLR^c \cdot (-1000) \\
\widehat{LE} = & \big(\widehat\beta_0 - 1000 \widehat\beta_2 \big)+ \\ & \big(\widehat\beta_1 - 1000 \widehat\beta_3  \big) FLR^c
\end{aligned}$$
:::
:::
:::

::: {.column width="33%"}
::: FS2
::: FS2-title
Food Supply of `r mean_FS` kcal PPD
:::

::: FS2-cont
$$\begin{aligned} 
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \\ & \widehat\beta_2 \cdot 0 + \\ & \widehat\beta_3 FLR^c \cdot 0 \\
\widehat{LE} = & \big(\widehat\beta_0 \big)+ \\ & \big(\widehat\beta_1 \big) FLR^c
\end{aligned}$$
:::
:::
:::

::: {.column width="33%"}
::: FS3
::: FS3-title
Food Supply of `r mean_FS + 1000` kcal PPD
:::

::: FS3-cont
$$\begin{aligned} 
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \\ & \widehat\beta_2 \cdot 1000 + \\ & \widehat\beta_3 FLR^c \cdot 1000 \\
\widehat{LE} = & \big(\widehat\beta_0 + 1000 \widehat\beta_2 \big)+ \\ & \big(\widehat\beta_1 + 1000 \widehat\beta_3  \big) FLR^c
\end{aligned}$$
:::
:::
:::
:::

## Poll Everywhere Question??

## Interpretation for interaction between two continuous variables

$$ \begin{aligned}
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \widehat\beta_2 FS^c + \widehat\beta_3 FLR^c \cdot FS^c \\
\widehat{LE} = & \bigg[\widehat\beta_0 + \widehat\beta_2 \cdot FS^c \bigg] + \underbrace{\bigg[\widehat\beta_1 + \widehat\beta_3 \cdot FS^c \bigg]}_\text{FLR's effect} FLR \\
\end{aligned}$$

-   Interpretation:

    -   $\beta_3$ = mean change in female literacy rate's effect, for every one kcal PPD increase in food supply

-   In summary, the interaction term can be interpreted as "difference in adjusted female literacy rate effect for every 1 kcal PPD increase in food supply"

-   It will be helpful to test the interaction to round out this interpretation!!

## Test interaction between two continuous variables

-   We run an F-test for a single coefficients ($\beta_3$) in the below model (see lesson 9)

$$ LE = \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \beta_3 FLR^c \cdot FS^c + \epsilon$$

::: columns
::: {.column width="10%"}
:::

::: {.column width="40%"}
::: proof1
::: proof-title
Null $H_0$
:::

::: proof-cont
$$\beta_3=0$$
:::
:::
:::

::: {.column width="40%"}
::: definition
::: def-title
Alternative $H_1$
:::

::: def-cont
$$\beta_3\neq0$$
:::
:::
:::

::: {.column width="10%"}
:::
:::

::: columns
::: {.column width="10%"}
:::

::: {.column width="40%"}
::: proof1
::: proof-title
Null / Smaller / Reduced model
:::

::: proof-cont
$$ LE = \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \epsilon$$
:::
:::
:::

::: {.column width="40%"}
::: definition
::: def-title
Alternative / Larger / Full model
:::

::: def-cont
$$\begin{aligned}
LE = & \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \\ & \beta_3 FLR^c \cdot FS^c + \epsilon
\end{aligned}$$
:::
:::
:::
:::

## Test interaction between two continuous variables

-   Fit the reduced and full model

```{r}
#| echo: true
m_int_fs_red = lm(LifeExpectancyYrs ~ FLR_c + FS_c, 
                   data = gapm_sub)
m_int_fs_full = lm(LifeExpectancyYrs ~ FLR_c + FS_c +
                  FLR_c*FS_c, data = gapm_sub)
```

 

```{r}
#| code-fold: true
#| code-summary: "Display the ANOVA table with F-statistic and p-value"
#| echo: true
anova(m_int_fs_red, m_int_fs_full) %>% tidy() %>% 
  gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 3)
```

-   Conclusion: There is not a significant interaction between female literacy rate and food supply (p = 0.945). Food supply is not an effect modifier of the association between female literacy rate and life expectancy.

# Learning Objective

5.  Interpret the interaction component of a model with **two continuous covariates**, and how the main variable's effect changes.

::: lob
6.  When there are only two covariates in the model, test whether one is a confounder or effect modifier.
:::

## Deciding between confounder and effect modifier

-   This is more of a model selection question (in coming lectures)

-   But if we had a model with **only TWO covariates**, we could step through the following process:

    1.  Test the interaction (of potential effect modifier): use a partial F-test to test if interaction term(s) explain enough variation compared to model without interaction

        -   Recall that for two continuous covariates, we will **test a single coefficient**
        -   For a binary and continuous covariate, we will **test a single coefficient**
        -   For two binary categorical covariates, we will **test a single coefficient**
        -   **For a multi-level categorical covariate (with any other type of covariate), we must test a group of coefficients!!**

    2.  Then look at the main effect (or potential confounder)

        -   If interaction already included, then automatically included as main effect (and thus not checked for confounding)

        -   For variables that are not included in any interactions:

            -   Check to see if they are confounders by seeing whether exclusion of the variable changes any of the main effect of the primary explanatory variable by more than 10%

## Reminder from Lesson 9: General steps for F-test

::: columns
::: {.column width="48%"}
::: highlight-container
::: highlight
1.  Met underlying LINE assumptions
:::
:::

::: highlight-container
::: highlight
2.  State the null hypothesis
:::
:::

```{=tex}
\begin{align}
H_0 &: \beta_1=\beta_2= \ldots=\beta_k=0\\
\text{vs. } H_A&: \text{At least one } \beta_j\neq0, \text{for }j=1, 2, \ldots, k
\end{align}
```
::: highlight-container
::: highlight
3.  Specify the significance level.
:::
:::

Often we use $\alpha = 0.05$

::: highlight-container
::: highlight
4.  Specify the test statistic and its distribution under the null
:::
:::

The test statistic is $F$, and follows an F-distribution with numerator $df=k$ and denominator $df=n-k-1$. ($n$ = \# obversation, $k$ = \# covariates)
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
::: highlight-container
::: highlight
5.  Compute the value of the test statistic
:::
:::

The calculated **test statistic** is

$$F^ = \dfrac{\frac{SSE(R) - SSE(F)}{df_R - df_F}}{\frac{SSE(F)}{df_F}} = \frac{MSR_{full}}{MSE_{full}}$$

::: highlight-container
::: highlight
6.  Calculate the p-value
:::
:::

We are generally calculating: $P(F_{k, n-k-1} > F)$

::: highlight-container
::: highlight
7.  Write conclusion for hypothesis test
:::
:::

We (reject/fail to reject) the null hypothesis at the $100\alpha\%$ significance level. <Include specific context for the test>
:::
:::

## Step 1: Testing the interaction

-   We test with $\alpha = 0.10$

-   Follow the F-test procedure in [Lesson 9 (MLR: Inference/F-test)](.\slides\09_MLR_Inference.qmd)

    -   This means we need to follow the 7 steps of the general F-test in previous slide (taken from Lesson 9)

-   Use the hypothesis tests for the specific variable combo:

::: columns
::: column
::: fact
::: fact-title
Binary & continuous variable (Lesson 11, LOB 2)
:::

::: fact-cont
Testing a single coefficient for the interaction term using F-test comparing full model to reduced model
:::
:::
:::

::: column
::: definition
::: def-title
Multi-level & continuous variables (Lesson 11, LOB 3)
:::

::: def-cont
Testing group of coefficients for the interaction terms using F-test comparing full to reduced model
:::
:::
:::
:::

::: columns
::: column
::: proposition
::: prop-title
Binary & multi-level variable (Lesson 11, LOB 4)
:::

::: prop-cont
Testing group of coefficients for the interaction terms using F-test comparing full to reduced model
:::
:::
:::

::: column
::: proof1
::: proof-title
Two continuous variables (Lesson 11, LOB 5)
:::

::: proof-cont
Testing a single coefficient for the interaction term using F-test comparing full to reduced model
:::
:::
:::
:::

## Poll Everywhere Questions 2-4

## Step 2: Testing a confounder

-   If interaction already included:

    -   Meaning: F-test showed evidence for alternative/full model
    -   Then the variable is an effect modifier and we don't need to consider it as a confounder
    -   Then automatically included as main effect (and thus not checked for confounding)

-   For variables that are not included in any interactions:

    -   Check to see if they are confounders
    -   One way to do this is by seeing whether exclusion of the variable changes any of the main effect of the primary explanatory variable by more than 10%

-   If the main effect of the primary explanatory variable changes by less than 10%, then the additional variable is neither an effect modifier nor a confounder

    -   We leave the variable out of the model

## Testing for percent change ( $\Delta\%$) in a coefficient

-   Let's say we have $X_1$ and $X_2$, and we specifically want to see if $X_2$ is a confounder for $X_1$ (the explanatory variable or variable of interest)

-   If we are only considering $X_1$ and $X_2$, then we need to run the following two models:

    -   Fitted model 1 / reduced model (`mod1`): $\widehat{Y} = \widehat\beta_0 + \widehat\beta_1X_1$

        -   We call the above $\widehat\beta_1$ the reduced model coefficient: $\widehat\beta_{1, \text{mod1}}$ or $\widehat\beta_{1, \text{red}}$

    -   Fitted model 2 / Full model (`mod2`): $\widehat{Y} = \widehat\beta_0 + \widehat\beta_1X_1 +\widehat\beta_2X_2$

        -   We call this $\widehat\beta_1$ the full model coefficient: $\widehat\beta_{1, \text{mod2}}$ or $\widehat\beta_{1, \text{full}}$

::: columns
::: {.column width="15%"}
:::

::: {.column width="70%"}
::: fact
::: fact-title
Calculation for % change in coefficient
:::

::: fact-cont
$$
\Delta\% = 100\% \cdot\frac{\widehat\beta_{1, \text{mod1}} - \widehat\beta_{1, \text{mod2}}}{\widehat\beta_{1, \text{mod2}}} = 100\% \cdot \frac{\widehat\beta_{1, \text{red}} - \widehat\beta_{1, \text{full}}}{\widehat\beta_{1, \text{full}}}
$$
:::
:::
:::

::: {.column width="15%"}
:::
:::

## Is food supply a confounder for female literacy rate? (1/3)

1.  Run models with and without food supply:

     

    -   Model 1 (reduced): $LE = \beta_0 + \beta_1 FLR^c + \epsilon$

    ```{r}
    #| echo: true
    mod1_red = lm(LifeExpectancyYrs ~ FLR_c, data = gapm_sub)
    ```

     

    -   Model 2 (full): $LE = \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \epsilon$

    ```{r}
    #| echo: true
    mod2_full = lm(LifeExpectancyYrs ~ FLR_c + FS_c, data = gapm_sub)
    ```

 

-   Note that the full model when testing for confounding was the reduced model for testing an interaction

-   Full and reduced are always relative qualifiers of the models that we are testing

## Is food supply a confounder for female literacy rate? (2/3)

2.  Record the coefficient estimate for centered female literacy rate in both models:

-   Model 1 (reduced):

```{r}
tidy(mod1_red, conf.int=T) %>% gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 5)
```

-   Model 2 (full):

```{r}
tidy(mod2_full, conf.int=T) %>% gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 5)
```

3.  Calculate the percent change:

$$
\Delta\% = 100\% \cdot \frac{\widehat\beta_{1, \text{mod1}} - \widehat\beta_{1, \text{mod2}}}{\widehat\beta_{1, \text{mod2}}} = 100\% \cdot \frac{0.22990 - 0.15670}{0.15670} = `r round(100*(0.22990 - 0.15670)/0.15670, 2)`\%
$$

## Is food supply a confounder for female literacy rate? (3/3)

The percent change in female literacy rate's coefficient estimate was 46.71%.

Thus, food supply is a confounder of female literacy rate in the association between life expectancy and female literacy rate.

## Let's try this out on one of our potential effect modifiers or confounders

::: columns
::: column
-   Look back at income level and world region: is income level an **effect modifier, confounder, or has no effect** on the association between life expectancy and world region?

-   We can start by visualizing the relationship between life expectancy and world region *by income level*

-   So we'll need to revisit the work we did in previous slides on the interaction, then check fo condounding
:::

::: column
```{r fig.height=7, fig.width=9, warning=F, fig.align='center'}
#| echo: false
ggplot(gapm_sub, aes(x = four_regions, y = LifeExpectancyYrs, 
                  color = income_levels2)) +
  geom_jitter(size = 2, alpha = .6, width = 0.2) +
  stat_summary(fun = mean, geom = "point", size = 8, shape = 18) +
  labs(x = "World Regions", 
       y = "Life expectancy (years)",
       title = "Life expectancy vs. World Region", 
       color = "Income level") +
  theme(axis.title = element_text(size = 20), 
        axis.text = element_text(size = 20), 
        title = element_text(size = 20), 
        legend.text=element_text(size= 18)) +
  scale_color_manual(values=c("#FF8021", "#34AC8B"))
```
:::
:::

## Determining if income level is an **effect modifier, confounder, or neither**

-   Step 1: Testing the interaction/effect modifier

    -   Compare model with and without interaction using F-test to see if interaction is significant

    -   Models

        -   Model 1 (red): $\begin{aligned}LE = &\beta_0 + \beta_1 I(\text{Americas}) + \beta_2 I(\text{Asia}) + \beta_3 I(\text{Europe}) + \beta_4 I(\text{high income})+ \epsilon \end{aligned}$
        
         
        
        -   Model 2 (full): $\begin{aligned}LE = &\beta_0 + \beta_1 I(\text{Americas}) + \beta_2 I(\text{Asia}) + \beta_3 I(\text{Europe}) + \beta_4 I(\text{high income})+ \\ & \beta_5 \cdot I(\text{high income}) \cdot I(\text{Americas}) + \beta_6\cdot I(\text{high income}) \cdot I(\text{Asia})+ \\ &  \beta_7 \cdot I(\text{high income})\cdot I(\text{Europe})+ \epsilon \end{aligned}$

-   Step 2: Testing a confounder (only if not an effect modifier)

    -   Compare model with and without main effect for additional variable (income level) using F-test to see if additional variable (income level) is a confounder

    -   Models

        -   Model 1 (reduced): $\begin{aligned}LE = &\beta_0 + \beta_1 I(\text{Americas}) + \beta_2 I(\text{Asia}) + \beta_3 I(\text{Europe}) + \epsilon \end{aligned}$
        -   Model 2 (full): $\begin{aligned}LE = &\beta_0 + \beta_1 I(\text{Americas}) + \beta_2 I(\text{Asia}) + \beta_3 I(\text{Europe}) + \beta_4 I(\text{high income})+ \epsilon \end{aligned}$

## Step 1: Results from Lesson 11 LOB 4

-   Fit the reduced and full model

```{r}
#| echo: true
m_int_wr_inc_red = lm(LifeExpectancyYrs ~ income_levels2 + four_regions, 
                   data = gapm_sub)
m_int_wr_inc_full = lm(LifeExpectancyYrs ~ income_levels2 + four_regions +
                          income_levels2*four_regions, data = gapm_sub)
```

-   Display the ANOVA table with F-statistic and p-value

```{r}
anova(m_int_wr_inc_red, m_int_wr_inc_full) %>% tidy() %>% gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 3)

# newdata <- data.frame(four_regions = c("Africa", "Americas", "Asia", "Europe"), 
#                       FLR_c = c()) 
# (pred = predict(m_int_wr_full, 
#                 newdata=newdata, 
#                 interval="confidence"))
```

-   Conclusion: There is not a significant interaction between world region and income level (p = 0.928).

-   Thus, income level is not an effect modifier of world region. However, we can continue to test if income level is a confounder.

## Step 2: See if income is a confounder

-   Fit the reduced and full model for testing the confounder

 

-   Model 1 (reduced): $\begin{aligned}LE = &\beta_0 + \beta_1 I(\text{Americas}) + \beta_2 I(\text{Asia}) + \beta_3 I(\text{Europe}) + \epsilon \end{aligned}$

```{r}
#| echo: true
mod1_wr_inc_red = lm(LifeExpectancyYrs ~ four_regions, 
                   data = gapm_sub)
```

 

-   Model 2 (full): $\begin{aligned}LE = &\beta_0 + \beta_1 I(\text{Americas}) + \beta_2 I(\text{Asia}) + \beta_3 I(\text{Europe}) + \beta_4 I(\text{high income})+ \epsilon \end{aligned}$

```{r}
#| echo: true
mod1_wr_inc_full = lm(LifeExpectancyYrs ~ four_regions + income_levels2, 
                   data = gapm_sub)
```


## Step 2: See if income is a confounder

-   Record the coefficient estimate for centered female literacy rate in both models:

-   Model 1 (reduced):$\begin{aligned}\widehat{LE} = &\widehat\beta_0 + \widehat\beta_1 I(\text{Americas}) + \widehat\beta_2 I(\text{Asia}) + \widehat\beta_3 I(\text{Europe}) \end{aligned}$

```{r}
tidy(mod1_wr_inc_red, conf.int=T) %>% gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 5)
```

-   Model 2 (full): $\begin{aligned}\widehat{LE} = &\widehat\beta_0 + \widehat\beta_1 I(\text{Americas}) + \widehat\beta_2 I(\text{Asia}) + \widehat\beta_3 I(\text{Europe}) + \widehat\beta_4 I(\text{high income})\end{aligned}$

```{r}
tidy(mod1_wr_inc_full, conf.int=T) %>% gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 5)
```

## Step 2: See if income is a confounder

-   Calculate the percent change for $\widehat\beta_1$: $$
    \Delta\%  = 100\% \cdot \frac{\widehat\beta_{1, \text{mod1}} - \widehat\beta_{1, \text{mod2}}}{\widehat\beta_{1, \text{mod2}}} = 100\% \cdot \frac{14.33000 - 12.04102}{12.04102} = `r round(100*(14.33000 - 12.04102)/12.04102, 2)`
    $$

-   Calculate the percent change for $\widehat\beta_2$: $$
    \Delta\%  = 100\% \cdot \frac{\widehat\beta_{2, \text{mod1}} - \widehat\beta_{2, \text{mod2}}}{\widehat\beta_{2, \text{mod2}}} = 100\% \cdot \frac{8.11824 - 7.77808}{7.77808} = `r round(100*(8.11824 - 7.77808)/7.77808, 2)`
    $$

-   Calculate the percent change for $\widehat\beta_3$: $$
    \Delta\%  = 100\% \cdot \frac{\widehat\beta_{3, \text{mod1}} - \widehat\beta_{3, \text{mod2}}}{\widehat\beta_{3, \text{mod2}}} = 100\% \cdot \frac{14.78217 - 12.51938}{12.51938} = `r round(100*(14.78217 - 12.51938)/12.51938, 2)`
    $$

-   Note that two of these % changes are greater than 10%, and one is less than 10%...

## Step 2: See if income is a confounder

::: columns
::: column
-   There is no set rule when we have more than one estimated coefficient that we examine for confoundeing

-   In this, I would consider

    -   The majority of coefficients (2/3 coefficients) changes more than 10%
    -   The change in coefficients for all three are in the same direction
    -   The plot of life expectancy vs world region by income level have a shift in mean life expectancy from lower to higher income level
    
-   Thus, I would conclude that income level is a confounder, so we would leave income level's main effect in the model
:::

::: column
```{r fig.height=7, fig.width=9, warning=F, fig.align='center'}
#| echo: false
ggplot(gapm_sub, aes(x = four_regions, y = LifeExpectancyYrs, 
                  color = income_levels2)) +
  geom_jitter(size = 2, alpha = .6, width = 0.2) +
  stat_summary(fun = mean, geom = "point", size = 8, shape = 18) +
  labs(x = "World Regions", 
       y = "Life expectancy (years)",
       title = "Life expectancy vs. World Region", 
       color = "Income level") +
  theme(axis.title = element_text(size = 20), 
        axis.text = element_text(size = 20), 
        title = element_text(size = 20), 
        legend.text=element_text(size= 18)) +
  scale_color_manual(values=c("#FF8021", "#34AC8B"))
```
:::
:::

## If you want extra practice

-   Try out this procedure to determine if a variable is an effect modifier or confounder or nothing on the other interactions we tested out in Lesson 11

# Extra Reference Material

## General interpretation of the interaction term (reference)

$E[Y\mid X_{1},X_{2} ]=\beta_0 + \underbrace{(\beta_1+\beta_3X_{2}) }_\text{$X_{1}$'s effect} X_{1}+ \underbrace{\beta_2X_{2}}_\text{$X_{2}$ held constant}$\
${\color{white}{E[Y\mid X_{1},X_{2} ]}}=\beta_0 + \underbrace{(\beta_2+\beta_3X_{1}) }_\text{$X_{2}$'s effect}X_{2} + \underbrace{\beta_1X_{1}}_\text{$X_{1}$ held constant}$

-   Interpretation:

    -   $\beta_3$ = mean change in $X_{1}$'s effect, per unit increase in $X_{2}$;\

    -   [$\beta_3$]{style="color: white"} = mean change in $X_{2}$'s effect, per unit increase in $X_{1}$;

    -   where the "$X_{1}$ effect" equals the change in $E[Y]$ per unit increase in $X_{1}$ with $X_{2}$ held constant, i.e. "adjusted $X_{1}$ effect"

-   In summary, the interaction term can be interpreted as "difference in adjusted $X_1$ (or $X_2$) effect per unit increase in $X_2$ (or $X_1$)"

## A glimpse at how interactions might be incorporated into model selection {.smaller}

1.  Identify outcome (Y) and primary explanatory (X) variables

2.  Decide which other variables might be important and could be potential confounders. Add these to the model.

    -   This is often done by indentifying variables that previous research deemed important, or researchers believe could be important
    -   From a statistical perspective, we often include variables that are significantly associated with the outcome (in their respective SLR)

3.  (Optional step) Test 3 way interactions

    -   This makes our model incredibly hard to interpret. Our class will not cover this!!
    -   We will skip to testing 2 way interactions

4.  Test 2 way interactions

    -   When testing a 2 way interaction, make sure the full and reduced models contain the main effects

    -   First test all the 2 way interactions together using a partial F-test (with $alpha = 0.10$)

        -   If this test not significant, do not test 2-way interactions individually
        -   If partial F-test is significant, then test each of the 2-way interactions

5.  Remaining main effects - to include of not to include?

    -   For variables that are included in any interactions, they will be automatically included as main effects and thus not checked for confounding

    -   For variables that are not included in any interactions:

        -   Check to see if they are confounders by seeing whether exclusion of the variable(s) changes any of the coefficient of the primary explanatory variable (including interactions) X by more than 10%

            -   If any of X's coefficients change when removing the potential confounder, then keep it in the model
