---
title: "Lesson 11: Interactions Continued"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#213c96"
date: "02/14/2024"
categories: ["Week 7"]
format: 
  revealjs:
    theme: [default, simple_NW.scss]
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Interactions 2
    html-math-method: mathjax
    highlight-style: ayu
execute:
  freeze: auto  # re-render only when source changes
---


```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(knitr)
library(broom)
library(rstatix)
library(gt)
library(readxl)
#----------
# new packages
# install.packages("describedata")
library(describedata) # gladder()
library(gridExtra)   # grid.arrange()
library(ggfortify)  # autoplot(model)
# New Day 6
library(gtsummary)

# New Day 7
library(plotly) # for plot_ly() command
library(GGally) # for ggpairs() command 
library(ggiraphExtra)   # for ggPredict() command

# Load the data - update code if the csv file is not in the same location on your computer
# If you need to download the file, please go to ur shared folder under Data > Slides
gapm <- read_excel("data/Gapminder_vars_2011.xlsx", 
                   na = "NA")  # important!!!! 


gapm_sub <- gapm %>% # called it gapm2_sub3 to be consistent with Day 7 notes
  drop_na(LifeExpectancyYrs, FemaleLiteracyRate, four_regions, FoodSupplykcPPD) %>%
  mutate(four_regions = factor(four_regions, 
                               levels = c("africa", "americas", 
                                          "asia", "europe"), 
                               labels = c("Africa", "Americas", 
                                          "Asia", "Europe"))) %>%
  rename(income_levels = `World bank, 4 income groups 2017`) %>%
  mutate(income_levels1 = factor(income_levels, 
                                levels = c("Low income", 
                                           "Lower middle income", 
                                           "Upper middle income", 
                                           "High income")), 
         income_levels2 = relevel(factor(income_levels, 
                                levels = c("Low income", 
                                           "Lower middle income", 
                                           "Upper middle income", 
                                           "High income"), 
                                labels = c("Lower income", "Lower income", 
                                            "Higher income", "Higher income")), 
                                ref = "Lower income"))

```

# Learning Objective

5.  Interpret the interaction component of a model with **two continuous covariates**, and how the main variable's effect changes.
6.  When there are only two covariates in the model, test whether one is a confounder or effect modifier.

## Do we think food supply is an effect modifier for female literacy rate?

::: columns
::: {.column width="40%"}
-   We can start by visualizing the relationship between life expectancy and female literacy rate *by food supply*
-   Questions of interest: Does the effect of female literacy rate on life expectancy differ depending on food supply?

    -   This is the same as: Is food supply is an effect modifier for female literacy rate? Is food supply an effect modifier of the association between life expectancy and female literacy rate?

-   Let's run an interaction model to see!
:::

::: {.column width="60%"}
```{r fig.height=4, fig.width=6, warning=F, fig.align='center'}
#| echo: false
m_int_fs = lm(LifeExpectancyYrs ~ FemaleLiteracyRate*FoodSupplykcPPD, data = gapm_sub)
(mr1_2d = ggPredict(m_int_fs, interactive = T) )
```
:::
:::

## Model with interaction between *two continuous variables*

Model we are fitting:

$$ LE = \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \beta_3 FLR^c \cdot FS^c + \epsilon$$

-   $LE$ as life expectancy
-   $FLR^c$ as the **centered** around the mean female literacy rate (continuous variable)
-   $FS^c$ as the **centered** around the mean food supply (continuous variable)

```{r}
#| code-fold: true
#| code-summary: "Code to center FLR and FS"
#| echo: true
gapm_sub = gapm_sub %>% 
  mutate(FLR_c = FemaleLiteracyRate - mean(FemaleLiteracyRate), 
         FS_c = FoodSupplykcPPD - mean(FoodSupplykcPPD))
mean_FS = mean(gapm_sub$FoodSupplykcPPD) %>% round(digits = 0)
mean_FLR = mean(gapm_sub$FemaleLiteracyRate) %>% round(digits = 2)
```

In R: 
```{r}
#| echo: true

m_int_fs = lm(LifeExpectancyYrs ~ FLR_c + FS_c + FLR_c*FS_c, data = gapm_sub)
```

OR

```{r}
#| echo: true

m_int_fs = lm(LifeExpectancyYrs ~ FLR_c*FS_c, data = gapm_sub)
```

## Displaying the regression table and writing fitted regression equation

```{r}
#| echo: true
tidy_m_fs = tidy(m_int_fs, conf.int=T) 
tidy_m_fs %>% gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 5)
```

$$ \begin{aligned}
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \widehat\beta_2 FS^c + \widehat\beta_3 FLR^c \cdot FS^c \\
\widehat{LE} = & `r round(tidy_m_fs$estimate[1],2)` + `r round(tidy_m_fs$estimate[2],2)` \cdot FLR^c + `r round(tidy_m_fs$estimate[3],2)` \cdot FS^c - 0.00001 \cdot FLR^c \cdot FS^c
\end{aligned}$$

## Comparing fitted regression lines for various food supply values

$$ \begin{aligned}
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \widehat\beta_2 FS^c + \widehat\beta_3 FLR^c \cdot FS^c \\
\widehat{LE} = & `r round(tidy_m_fs$estimate[1],2)` + `r round(tidy_m_fs$estimate[2],2)` \cdot FLR^c + `r round(tidy_m_fs$estimate[3],2)` \cdot FS^c - 0.00001 \cdot FLR^c \cdot FS^c
\end{aligned}$$

To identify different lines, we need to pick example values of Food Supply:


::: columns
::: {.column width="33%"}
::: FS1
::: FS1-title
Food Supply of `r mean_FS -1000` kcal PPD
:::

::: FS1-cont
$$\begin{aligned} 
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \\ & \widehat\beta_2 \cdot (-1000) + \\ & \widehat\beta_3 FLR^c \cdot (-1000) \\
\widehat{LE} = & \big(\widehat\beta_0 - 1000 \widehat\beta_2 \big)+ \\ & \big(\widehat\beta_1 - 1000 \widehat\beta_3  \big) FLR^c
\end{aligned}$$
:::
:::
:::

::: {.column width="33%"}
::: FS2
::: FS2-title
Food Supply of `r mean_FS` kcal PPD
:::

::: FS2-cont
$$\begin{aligned} 
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \\ & \widehat\beta_2 \cdot 0 + \\ & \widehat\beta_3 FLR^c \cdot 0 \\
\widehat{LE} = & \big(\widehat\beta_0 \big)+ \\ & \big(\widehat\beta_1 \big) FLR^c
\end{aligned}$$
:::
:::
:::

::: {.column width="33%"}
::: FS3
::: FS3-title
Food Supply of `r mean_FS + 1000` kcal PPD
:::

::: FS3-cont
$$\begin{aligned} 
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \\ & \widehat\beta_2 \cdot 1000 + \\ & \widehat\beta_3 FLR^c \cdot 1000 \\
\widehat{LE} = & \big(\widehat\beta_0 + 1000 \widehat\beta_2 \big)+ \\ & \big(\widehat\beta_1 + 1000 \widehat\beta_3  \big) FLR^c
\end{aligned}$$
:::
:::
:::

:::

## Poll Everywhere Question??

## Interpretation for interaction between binary categorical and continuous variables

$$ \begin{aligned}
\widehat{LE} = & \widehat\beta_0 + \widehat\beta_1 FLR^c + \widehat\beta_2 FS^c + \widehat\beta_3 FLR^c \cdot FS^c \\
\widehat{LE} = & \bigg[\widehat\beta_0 + \widehat\beta_2 \cdot FS^c \bigg] + \underbrace{\bigg[\widehat\beta_1 + \widehat\beta_3 \cdot FS^c \bigg]}_\text{FLR's effect} FLR \\
\end{aligned}$$

-   Interpretation:

    -   $\beta_3$ = mean change in female literacy rate's effect, for every one kcal PPD increase in food supply

-   In summary, the interaction term can be interpreted as "difference in adjusted female literacy rate effect for every 1 kcal PPD increase in food supply"

-   It will be helpful to test the interaction to round out this interpretation!!

## Test interaction between multi-level categorical & continuous variables

-   We run an F-test for a group of coefficients ($\beta_5$, $\beta_6$, $\beta_7$) in the below model (see lesson 9)

$$ LE = \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \beta_3 FLR^c \cdot FS^c + \epsilon$$

::: columns

::: {.column width="10%"}
::: 

::: {.column width="40%"}
::: proof1
::: proof-title
Null $H_0$
:::

::: proof-cont
$$\beta_3=0$$
:::
:::
:::

::: {.column width="40%"}
::: definition
::: def-title
Alternative $H_1$
:::

::: def-cont
$$\beta_3\neq0$$ 
:::
:::
:::

::: {.column width="10%"}
:::

:::

::: columns

::: {.column width="10%"}
:::

::: {.column width="40%"}
::: proof1
::: proof-title
Null / Smaller / Reduced model
:::

::: proof-cont
$$ LE = \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \epsilon$$
:::
:::
:::

::: {.column width="40%"}
::: definition
::: def-title
Alternative / Larger / Full model
:::

::: def-cont
$$\begin{aligned}
LE = & \beta_0 + \beta_1 FLR^c + \beta_2 FS^c + \\ & \beta_3 FLR^c \cdot FS^c + \epsilon
\end{aligned}$$
:::
:::
:::
:::

## Test interaction between multi-level categorical & continuous variables

-   Fit the reduced and full model

```{r}
#| echo: true
m_int_fs_red = lm(LifeExpectancyYrs ~ FLR_c + FS_c, 
                   data = gapm_sub)
m_int_fs_full = lm(LifeExpectancyYrs ~ FLR_c + FS_c +
                  FLR_c*FS_c, data = gapm_sub)
```

 

```{r}
#| code-fold: true
#| code-summary: "Display the ANOVA table with F-statistic and p-value"
#| echo: true
anova(m_int_fs_red, m_int_fs_full) %>% tidy() %>% 
  gt() %>% tab_options(table.font.size = 35) %>% fmt_number(decimals = 3)
```

-   Conclusion: There is not a significant interaction between female literacy rate and food supply (p = 0.945). Food supply is not an effect modifier of the association between female literacy rate and life expectancy.

# Learning Objective

5.  Interpret the interaction component of a model with **two continuous covariates**, and how the main variable's effect changes.
6.  When there are only two covariates in the model, test whether one is a confounder or effect modifier.

## Deciding between confounder and effect modifier

-   This is more of a model selection question (in coming lectures)

-   But if we had a model with **only TWO covariates**, we could step through the following process:

    1.  Test the interaction (of potential effect modifier): use a partial F-test to test if interaction term(s) explain enough variation compared to model without interaction

        -   Recall that for two continuous covariates, we will **test a single coefficient**
        -   For a binary and continuous covariate, we will **test a single coefficient**
        -   For two binary categorical covariates, we will **test a single coefficient**
        -   **For a multi-level categorical covariate (with any other type of covariate), we must test a group of coefficients!!**

    2.  Then look at the main effect (or potential confounder)

        -   If interaction already included, then automatically included as main effect (and thus not checked for confounding)

        -   For variables that are not included in any interactions:

            -   Check to see if they are confounders by seeing whether exclusion of the variable changes any of the main effect of the primary explanatory variable by more than 10%

## Step 1: Testing the interaction

-   We test with $\alpha = 0.10$
-   Follow the F-test procedure

## Step 2: Testing a confounder

# Extra Reference Material

## General interpretation of the interaction term (reference)

$E[Y\mid X_{1},X_{2} ]=\beta_0 + \underbrace{(\beta_1+\beta_3X_{2}) }_\text{$X_{1}$'s effect} X_{1}+ \underbrace{\beta_2X_{2}}_\text{$X_{2}$ held constant}$\
${\color{white}{E[Y\mid X_{1},X_{2} ]}}=\beta_0 + \underbrace{(\beta_2+\beta_3X_{1}) }_\text{$X_{2}$'s effect}X_{2} + \underbrace{\beta_1X_{1}}_\text{$X_{1}$ held constant}$

-   Interpretation:

    -   $\beta_3$ = mean change in $X_{1}$'s effect, per unit increase in $X_{2}$;\

    -   [$\beta_3$]{style="color: white"} = mean change in $X_{2}$'s effect, per unit increase in $X_{1}$;

    -   where the "$X_{1}$ effect" equals the change in $E[Y]$ per unit increase in $X_{1}$ with $X_{2}$ held constant, i.e. "adjusted $X_{1}$ effect"

-   In summary, the interaction term can be interpreted as "difference in adjusted $X_1$ (or $X_2$) effect per unit increase in $X_2$ (or $X_1$)"

## A glimpse at how interactions might be incorporated into model selection {.smaller}

1.  Identify outcome (Y) and primary explanatory (X) variables

2.  Decide which other variables might be important and could be potential confounders. Add these to the model.

    -   This is often done by indentifying variables that previous research deemed important, or researchers believe could be important
    -   From a statistical perspective, we often include variables that are significantly associated with the outcome (in their respective SLR)

3.  (Optional step) Test 3 way interactions

    -   This makes our model incredibly hard to interpret. Our class will not cover this!!
    -   We will skip to testing 2 way interactions

4.  Test 2 way interactions

    -   When testing a 2 way interaction, make sure the full and reduced models contain the main effects

    -   First test all the 2 way interactions together using a partial F-test (with $alpha = 0.10$)

        -   If this test not significant, do not test 2-way interactions individually
        -   If partial F-test is significant, then test each of the 2-way interactions

5.  Remaining main effects - to include of not to include?

    -   For variables that are included in any interactions, they will be automatically included as main effects and thus not checked for confounding

    -   For variables that are not included in any interactions:

        -   Check to see if they are confounders by seeing whether exclusion of the variable(s) changes any of the coefficient of the primary explanatory variable (including interactions) X by more than 10%

            -   If any of X's coefficients change when removing the potential confounder, then keep it in the model
