---
title: "Lesson 12: In-class exercise!!"
subtitle: "Choosing what to test, interpretations of coefficients, F-test conclusions, and interactions"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#213c96"
date: "02/28/2024"
categories: ["Week 8"]
format: 
  revealjs:
    theme: [default, simple_NW.scss]
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: Model Selection
    html-math-method: mathjax
    highlight-style: ayu
execute:
  freeze: auto  # re-render only when source changes
---

\

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(knitr)
library(broom)
library(rstatix)
library(gt)
library(readxl)
library(describedata) # gladder()
library(gridExtra)   # grid.arrange()
library(ggfortify)  # autoplot(model)
library(gtsummary)
library(plotly) # for plot_ly() command
library(GGally) # for ggpairs() command 
library(ggiraphExtra)   # for ggPredict() command

SBP = read_excel("~/Library/CloudStorage/OneDrive-OregonHealth&ScienceUniversity/Teaching/Resources_from_Meike/B512_W23_shared/textbook/datasets/BSTA512_datasets/CH11/CH11Q05.xls") %>%
  rename(SBP = Y, PAL = X1, AGE = X2, SAB = X3)

# SBP = read_excel("~/Library/CloudStorage/OneDrive-OregonHealth&ScienceUniversity/Teaching/Resources_from_Meike/B512_W23_shared/textbook/datasets/BSTA512_datasets/CH05/CH05Q02.xls")

voter_data <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv")

library(palmerpenguins)
data(penguins)
```

## Example 1: Palmer Penguins

-   Revisit the Palmer Penguins dataset that we say in HW 4

## Let's take a look at the variables

```{r}
glimpse(penguins)
```

## Questions we can ask and answer so far...

-   Using SLR, does each variable predict flipper length significantly?

-   

## Example

-   The following example comes from **this textbook**

> Suppose that we are conducting an observational study of adults to assess whether physical activity level (PAL) is associated with systolic blood pressure (SBP), accounting (i.e., controlling) for AGE and SAB (sex assigned at birth). The extraneous variable here is AGE and SAB, while the explanatory variable (variable of interest) is PAL. We need to determine whether we can ignore AGE and/or SAB in our analysis and still correctly assess the PAL–SBP association.

## Explore the data: First thing to do

```{r}
#| code-fold: true

ggplot(SBP) + geom_histogram(aes(x = SBP))
ggplot(SBP) + geom_histogram(aes(x = PAL))
ggplot(SBP) + geom_histogram(aes(x = AGE))
```
