---
title: "Module_I"
author: Xu Shi
date: Oct 24, 2020
fontsize: 10pt
output: 
    beamer_presentation
latexengine: xelatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 70)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
## Example: 
+ We return to the data set containing age (measured in days) and birth weight (BWT in ounces) as predictors of infant systolic blood pressure (SBP, mm Hg). 
+ Read in the data file (systolic1.csv). (n = 16)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data = read.csv(file="systolic1.csv",header=T)
attach(data)
head(data)
```

## Fit SBP on AGE x BWT, AGE, BWT in this order
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m1 = lm(sbp~I(age*bwt)+I(age)+I(bwt))
summary(m1)
```

## Compute the sequential and partial SS
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
( SeqSS.m1 = anova(m1) ) #type I
( ParSS.m1 = car::Anova(m1,type="III") ) #type III
```


## You have two F tests of the interaction. Are either of them valid tests?


## Change ordering of covariates and recompute SS
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
( anova(lm(sbp~I(age)+I(bwt)+I(age*bwt))) ) #type I
( car::Anova(m1,type="III") ) #type III
```


## Save the fitted values
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data = cbind(data,fit.m1 = m1$fitted.values)
summary(data)
```

## Hypothesis test of interaction term
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~age*bwt))$coef
anova(lm(sbp~age*bwt))
```

## Hypothesis test of age
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
null.model = lm(sbp~bwt)
alt.model = lm(sbp~age*bwt)
anova(null.model,alt.model)
```

## Compare to the adjusted age effect in an *additive* model 
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
# car::Anova(lm(sbp~age),type="III") #crude effect
anova(lm(sbp~bwt),lm(sbp~age+bwt))  #adjusted effect in additive model
car::Anova(lm(sbp~age+bwt),type="III") #adjusted effect in additive model
```

## Hypothesis test of age: by hand
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
SeqSS = data.frame(anova(lm(sbp~I(bwt)+I(age)+I(bwt*age))))
df1 = SeqSS["I(age)","Df"]+SeqSS["I(bwt * age)","Df"]
F1 = (SeqSS["I(age)","Sum.Sq"]+SeqSS["I(bwt * age)","Sum.Sq"])/df1
df2 = SeqSS["Residuals","Df"]
F2 = SeqSS["Residuals","Mean.Sq"]
c(F.val = F1/F2,
  P.val = 1 - pf(F1/F2,df1=df1,df2=df2))
anova(null.model,alt.model) ##validate
```

## Hypothesis test of bwt
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
null.model = lm(sbp~age)
alt.model = lm(sbp~age*bwt)
anova(null.model,alt.model)
```



## Center the covariates
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
c(mean(age), sd(age))
c(mean(bwt), sd(bwt))
age.c = age - mean(age)
bwt.c = bwt - mean(bwt)
```

## Fit SBP on AGE x BWT, AGE, BWT, this time using the centered versions of the covariates.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m2 = lm(sbp~I(age.c*bwt.c)+I(age.c)+I(bwt.c))
summary(m2)
```

## Compute the sequential and partial SS
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
( SeqSS.m2 = anova(m2) ) #type I
( ParSS.m2 = car::Anova(m2,type="III") ) #type III
```

## Save the fitted values
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data = cbind(data,fit.m2 = m2$fitted.values)
summary(data)
```

## Compare the output from m1 and m2, what is the same? what is different? 
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(m1)$coef
summary(m2)$coef
```

## Compare the output from m1 and m2, what is the same? what is different? 
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
rbind(data.frame(SeqSS.m1),data.frame(SeqSS.m2))
rbind(data.frame(ParSS.m1)[-1,],data.frame(ParSS.m2)[-1,])
```

## Verify the models are numerically equivalent
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(data)
```

## Interpretation in uncentered and centered models.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~age*bwt))$coef
summary(lm(sbp~age.c*bwt.c))$coef
```
+ (1) Give interpretations for the main effect coefficient for AGE 
+ (2) Interpret the interaction


## (1) Give interpretations for the main effect coefficient for AGE 

## (2) Interpret the interaction

## Dichotomize BWT. Refit the interaction model.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
highbwt = 1*(bwt > mean(bwt)) #1*(bwt.c > 0)
mgrp = lm(sbp~age*highbwt)
summary(mgrp)
```

## Scatterplot with subjects stratified by age category
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
plot(age[highbwt == 0], sbp[highbwt == 0], pch = 15, col="black",
     xlab = "Age (years)", ylab = "SBP (mm Hg)",xlim=range(age),ylim=range(sbp))
points(age[highbwt == 1], sbp[highbwt == 1], pch = 16, col="red")
lines(age[highbwt == 0],mgrp$fitted.values[highbwt == 0],col="black")
lines(age[highbwt == 1],mgrp$fitted.values[highbwt == 1],col="red")
legend("bottomright",bty="n",legend=c("higher bwt","lower bwt"),
       col=c("red","black"),pch=c(16,15))
```


## Fit regression by subgroup
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
mgrp1 = lm(sbp~age,data=data.frame(sbp,age)[highbwt==1,])
mgrp0 = lm(sbp~age,data=data.frame(sbp,age)[highbwt==0,])
summary(mgrp)$coef
summary(mgrp1)$coef
summary(mgrp0)$coef
```

## Fit regression by subgroup
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
rbind( mgrp0=coef(mgrp0), mgrp=coef(mgrp)[c("(Intercept)","age")] )
rbind( mgrp1=coef(mgrp1), 
   mgrp=c(
   coef(mgrp)["(Intercept)"]+coef(mgrp)["highbwt"],
   coef(mgrp)["age"]+coef(mgrp)["age:highbwt"]
   )
)
```


## Get 95% CI for age effect among high BWT subjects.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
mgrp = lm(sbp~age*highbwt)
vcov(mgrp)
age.eff = c(  c(0,1,0,1)%*%coef(mgrp)  )
age.var = c(  c(0,1,0,1)%*%vcov(mgrp)%*%c(0,1,0,1)  )
t = qt(p=0.975,df=nrow(data)-4) #or qnorm(0.975) if n=nrow(data) is large
(age.confint = age.eff + c(-1,1)*t*sqrt(age.var))
confint(mgrp1)
```


