---
title: "Module_I"
author: "Nicky Wakim"
date: "10/30/2023"
fontsize: 10pt
output: 
    beamer_presentation
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 70)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
## Example: 
+ Read in the data file (hdl.csv). (n = 26)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data = read.csv(file="hdl.csv",header=T)
attach(data)
head(data)
```

## The MLR model
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
G1 = as.numeric(group==1)
G2 = as.numeric(group==2)
G3 = as.numeric(group==3)
wt = weight - mean(weight)
m1 = lm(hdl~-1+I(G1)+I(G2)+I(G3)+I(G1*wt)+I(G2*wt)+I(G3*wt))
# alternatively: lm(hdl~-1+factor(group)*wt-wt)
summary(m1)$coef
```

## GLH test for parallelism
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(Contrast.T = matrix(c(0,0,0,1,-1,0,0,0,0,0,1,-1),byrow=T,nrow=2))
car::linearHypothesis(model=m1,hypothesis.matrix=Contrast.T,rhs=c(0,0))
```


## GLH test for equality of intercepts
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(Contrast.T = matrix(c(1,-1,0,0,0,0,0,1,-1,0,0,0),byrow=T,nrow=2))
car::linearHypothesis(model=m1,hypothesis.matrix=Contrast.T,rhs=c(0,0))
```


## The MLR model reparameterized
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m2 = lm(hdl~I(G2)+I(G3)+I(wt)+I(G2*wt)+I(G3*wt))
anova(m2) #Sequential SS (type I)
(pval = 1-pf(q = (317.97+187.09)/2/85.62, df1 = 2, df2 = 20))
```

## Change ordering of covariates
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m3 = lm(hdl~I(wt)+I(G2*wt)+I(G3*wt)+I(G2)+I(G3))
anova(m3) #Sequential SS (type I)
(pval = 1-pf(q = (509.73+25.24)/2/85.62, df1 = 2, df2 = 20))
```

## SS test Alternative 1: GLH formulation (slopes)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(Contrast.T = matrix(c(0,0,0,0,1,0,0,0,0,0,0,1),byrow=T,nrow=2))
car::linearHypothesis(model=m2,hypothesis.matrix=Contrast.T,rhs=c(0,0))
```

## SS test Alternative 1: GLH formulation (intercepts)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(Contrast.T = matrix(c(0,1,0,0,0,0,0,0,1,0,0,0),byrow=T,nrow=2))
car::linearHypothesis(model=m2,hypothesis.matrix=Contrast.T,rhs=c(0,0))
```

## SS test Alternative 2: ANOVA of nested models (slopes)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
anova(
  lm(hdl~I(G2)+I(G3)+I(wt)+I(G2*wt)+I(G3*wt)),
  lm(hdl~I(G2)+I(G3)+I(wt))
)
```

## SS test Alternative 2: ANOVA of nested models (intercepts)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
anova(
  lm(hdl~I(G2)+I(G3)+I(wt)+I(G2*wt)+I(G3*wt)),
  lm(hdl~I(wt)+I(G2*wt)+I(G3*wt))
)
```