---
title: "Module_H"
author: Xu Shi
date: Oct 18, 2020
fontsize: 10pt
output: 
    beamer_presentation
latexengine: xelatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 70)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
```
## Example: 
+ A data set was assembled consisting of life expectancy calculations, as well as other population level demographic data. Each record in the data set pertains to a country, with the primary objective being to determine the role of the following descriptors as predictors of life expectancy: continent, sex, persons per television, and persons per medical doctors.
+ Read in the data file (life1.csv).
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data = read.csv(file="life1.csv",header=T)
attach(data)
head(data)
```

## Produce frequency tables for continent.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
Hmisc::describe(continent)
```

## Compute mean life expectancy by continent.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
by(life_exp, factor(continent), mean) #tapply
```

## Create dummy variables for continent
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
dummy_Af = 1*(continent=="Af")
dummy_As = 1*(continent=="As")
dummy_Eu = 1*(continent=="Eu")
dummy_Na = 1*(continent=="Na")
dummy_Sa = 1*(continent=="Sa")
```


## Fit a linear model with all continents represented. Interpret the NA in the output.
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
m0 = lm(life_exp~dummy_Af+dummy_As+dummy_Eu+dummy_Na+dummy_Sa); summary(m0)
```

## Compute the fitted means for each continent and compare them to the means computed previously.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
coefs = c(coef(m0)) ## beta coefs
coefs["dummy_Sa"] = 0 ## set NA to 0
coefs[-1] + coefs["(Intercept)"] ## add intercept to each
tapply(life_exp, factor(continent), mean)
```

## Fit a model w/o an intercept. Compare the parameter estimates to the group means.
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
m1 = lm(life_exp~ -1+dummy_Af+dummy_As+dummy_Eu+dummy_Na+dummy_Sa); summary(m1)
# Af       As       Eu       Na       Sa 
# 59.08333 65.03571 74.15000 74.66667 69.10000 
```

## Compare the SSE values for the two regression models fitted thus far. How can they be equal when SSY is defined differently?
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
data.frame(anova(m0))
data.frame(anova(m1))
```


## Why does a linear model without an intercept (forced through the origin) have a higher R-squared value?
\pause
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
ones = rep(1,nrow(data))
m1.2 = lm(life_exp~ -1+ones+dummy_Af+dummy_As+dummy_Eu+dummy_Na+dummy_Sa); summary(m1.2)
```


## Interpret each row of the results in the no-intercept model.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(m1)
```

## Fit a model using the factor() function.
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
m2.1 = lm(life_exp~factor(continent)); summary(m2.1)
```

## Fit a model using dummy variables to get m2.2.
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
m2.2 = lm(life_exp~dummy_As+dummy_Eu+dummy_Na+dummy_Sa); summary(m2.2)
```

## Compare ANOVA tables of m2.1 and m2.2
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
data.frame(anova(m2.1))
data.frame(anova(m2.2))
sum(data.frame(anova(m2.2))[1:4,"Sum.Sq"])
```

## Interpret each row of the results in m2.1.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(m2.1)
```

## Fit a model using all covariates.
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
(m = lm(life_exp~sex+factor(continent)+per_TV+per_MD))
```
+ Interpret the intercept, and describe any related concerns.

## What percentage of the variation in life expectancy is explained by gender *alone*?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
( SSY = var(life_exp)*(nrow(data)-1) )
(SeqSS = data.frame(anova(lm(life_exp~sex))))
sum(SeqSS[,"Sum.Sq"])
SeqSS["sex","Sum.Sq"]/SSY
```

## What percentage of the variation in life expectancy is explained by gender *alone*?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
( SSY = var(life_exp)*(nrow(data)-1) )
(ParSS = data.frame(car::Anova(lm(life_exp~sex),type="III")))
sum(ParSS[-1,"Sum.Sq"])
ParSS["sex","Sum.Sq"]/SSY
```

## Consider the percentage of variation explained by the model with all covariates. What fraction of this percentage is attributable to the per-MD and per-TV covariates?
\pause
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
(SeqSS = data.frame(anova(m)))
(R2_MDTV = (SeqSS["per_TV","Sum.Sq"]+SeqSS["per_MD","Sum.Sq"])/SSY)
R2_MDTV/summary(m)$r.squared
(SeqSS["per_TV","Sum.Sq"]+SeqSS["per_MD","Sum.Sq"])/sum(SeqSS[1:4,"Sum.Sq"])
```