---
title: "Module_K"
author: Nicky
date: Nov 2, 2021
fontsize: 10pt
output: 
    beamer_presentation
latexengine: pdflatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 70)
options(warn=-1)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
## Example: 
+ A common measurement of lung function is the forced expiratory volume (FEV), which measures how much air you can blow out of your lungs in a short period of time. A higher FEV is usually associated with better respiratory function. 
A group of clinicians is interested in examining the factors which affect FEV1. Data on 654 children are available for the following covariates: AGE (A$_i$, years), HEIGHT (H$_i$, inches), MALE (M$_i$=0 for female, 1 for male) SMOKE (S$_i$=0 for non-smoker and 1 for smoker). Each row corresponds to a single clinic visit.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
rm(list=ls())
data = read.csv(file="fev.csv",header=T)
attach(data)
head(data)
```

## Compute descriptive statistics on the covariates. Any concerns?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
# psych::describe(data)
psych::pairs.panels(data)
```

## Fit a main effects model and obtain  $\widehat{\sigma}, \widehat{Y}_i,\widehat{\epsilon}_i,\widehat{r}_i,\widehat{r}_{(-i)}$, and $h_{ii}$
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m1 = lm(fev~A+H+M+S)
m1.yhat = m1$fitted.values
m1.res = m1$residuals #m1.yhat-fev
m1.h = hatvalues(m1)  #leverage
m1.r = rstandard(m1)  #internally studentized residuals
m1.rr = rstudent(m1)  #externally studentized residuals
```

## Which subject is most outlying with respect to the X space?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
Hmisc::describe(m1.h)
m1.h[which.max(m1.h)]
```

## Examine the quartiles of the $\widehat{\epsilon}_i$. What do you think?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
Hmisc::describe(m1.res)
Hmisc::describe(m1.res)$counts[c(".25",".50",".75")]
```


## Do a histogram of the residuals. What do you think?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
hist(m1.res,breaks=15)
```

## Do a Q/Q plot of the residuals. What do you think?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
qq.m1.res = car::qqPlot(m1.res)
m1.res[qq.m1.res]
```


## Compute the $\widehat{z}_i$ -- this is typically not provided in R or SAS because it is imperfect.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
n = nrow(data); p = m1$rank # dimensions
m1.sigma = sqrt(  sum(m1.res^2)/(n-p)  )
m1.z = m1.res/m1.sigma #standardized residual
```

## For what percentage are (i) $|\widehat{z}_i|>1$ (ii) $|\widehat{z}_i|>2$ (iii) $|\widehat{z}_i|>2.5$
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m1.z.g1 = as.numeric(abs(m1.z)>1); mean(m1.z.g1)*100
m1.z.g2 = as.numeric(abs(m1.z)>2); mean(m1.z.g2)*100
m1.z.g2.5 = as.numeric(abs(m1.z)>2.5); mean(m1.z.g2.5)*100
```



## Do partial regression plots. Do you have evidence of non-linear relationships?
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
car::avPlots(m1)
```


## Residual plots.
```{r mysize=T,size='\\tiny', echo=T,eval=T}
#car::residualPlots(m1,type="response")
```

## Fit a model which replaces HEIGHT with a set of quartiles. Do you have any evidence that the relationship between HEIGHT and FEV1 is non-linear?
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
H.grp = cut(H,breaks=quantile(H, probs=seq(0,1, by=0.25)),include.lowest=T)
m2 = lm(fev~A+H.grp+M+S)
summary(m2)
```

## Fit a model which replaces HEIGHT with a set of quartiles. Do you have any evidence that the relationship between HEIGHT and FEV1 is non-linear?
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
car::avPlots(m2)
```

### Residual plots.
```{r mysize=T,size='\\tiny', echo=T,eval=T}
#car::residualPlots(m2,type="response")
```

## Fit a model that contains a quadratic term for HEIGHT. Do you have any evidence of non-linearity?
```{r mysize=T,size='\\scriptsize', echo=T,eval=T}
H.c = H-median(H,na.rm=T)
m3 = lm(fev~A+H.c+I(H.c^2)+M+S)
summary(m3)
```

## Reconcile differences in your conclusions from m2 and m3.

## Compare plots of $\widehat{Y}_i$ vs $H_i$ based on each of the linear and quadratic models. Does the squared term appear to make a difference to the fitted values?
```{r mysize=T,size='\\footnotesize', echo=T,eval=F}
par(mfrow=c(2,3))
plot(x=H,y=m1$fitted.values,main="Yhat vs H in linear term model (m1)")
plot(x=fev,y=m1$fitted.values,main="Yhat vs Y in linear term model (m1)")
plot(x=H,y=m1$residuals,main="Residuals vs H in linear term model (m1)")
plot(x=H,y=m3$fitted.values,main="Yhat vs H in quadratic term model (m3)")
plot(x=fev,y=m3$fitted.values,main="Yhat vs Y in quadratic term model (m3)")
plot(x=H,y=m3$residuals,main="Residuals vs H in quadratic term model (m3)")
dev.off()
```

## Compare plots of $\widehat{Y}_i$ vs $H_i$ based on each of the linear and quadratic models. Does the squared term appear to make a difference to the fitted values?
```{r mysize=T,size='\\footnotesize', echo=F,eval=T}
par(mfrow=c(2,3))
plot(x=H,y=m1$fitted.values,main="Yhat vs H in linear term model (m1)")
plot(x=fev,y=m1$fitted.values,main="Yhat vs Y in linear term model (m1)")
plot(x=H,y=m1$residuals,main="Residuals vs H in linear term model (m1)")
plot(x=H,y=m3$fitted.values,main="Yhat vs H in quadratic term model (m3)")
plot(x=fev,y=m3$fitted.values,main="Yhat vs Y in quadratic term model (m3)")
plot(x=H,y=m3$residuals,main="Residuals vs H in quadratic term model (m3)")
# dev.off()
```
