---
title: "Module_N"
author: Xu Shi
date: Dec 1, 2020
fontsize: 10pt
output: 
    beamer_presentation
latexengine: pdflatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 70)
options(warn=-1)
```
```{r  mysize=T, tidy=T, size='\\footnotesize', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
## Example: 
+ A study was carried out on n=31 adult males in order to develop a prediction model for oxygen uptake. The study subjects run 1 mile on a treadmill, with an oxygen mask and heart monitor strapped to them. Data are recorded on age (years), weight (kg), time to run 1 mile (min), resting pulse (taken before the run), maximum pulse (during the run) and pulse at the end of the run; as well oxygen uptake.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
rm(list=ls())
data = read.csv(file="fitness1.csv",header=T)
attach(data)
head(data)
```

## Determine if there are any collinearity issues. Is any corrective action required?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
car::vif(lm(o2uptake~.,data=data))
```

## Generate a potential model using forward selection, with $p_I$ = 0.10.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_step_forward_p(lm(o2uptake~.,data=data),penter=0.1,details=F)
```


## If you re-run forward selection, but with $p_I$ = 0.05. How would your results compare with the previous ones?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_step_forward_p(lm(o2uptake~.,data=data),penter=0.05,details=F)
```

## Select a model through backward elimination, with $p_O$ = 0.10.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_step_backward_p(lm(o2uptake~.,data=data),prem=0.1,details=F)
```

## If you re-run backward selection, but with $p_O$ = 0.5. How would your results compare with the previous ones?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_step_backward_p(lm(o2uptake~.,data=data),prem=0.5,details=F)
```

## Categorize age for illustration
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data$age = cut(data$age,
               breaks=c(min(age)-1,median(age),max(age)+1)) 
table(data$age)
```

## Now with new age, select a model through forward selection, with $p_I$ = 0.10.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_step_forward_p(lm(o2uptake~.,data=data),penter=0.1,details=F)
```
```
