---
title: "Module_G"
author: Xu Shi
date: Oct 12, 2020
fontsize: 10pt
output: 
    beamer_presentation
latexengine: xelatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 65)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
```
## Example: 
+ A study is conducted to examine the role of age (measured in days) and birth weight (BWT in ounces) as predictors of infant systolic blood pressure (SBP, mm Hg). 
+ Read in the data file (systolic1.csv).
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data = read.csv(file="systolic1.csv",header=T)
attach(data)
```

## Estimate the crude (i.e., unadjusted) effect of BWT on SPB. Is the effect significant?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~bwt))
```

## Estimate the effect of BWT on SPB, adjusted for AGE. Is the effect significant?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~bwt+age))
```

## Estimate the correlation between BWT and AGE. Is the correlation significant?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
cor.test(bwt,age)
# summary(lm(bwt~age))$coef
# summary(lm(age~bwt))$coef
```

## Compare the simple and multiple linear regression coefficient estimates for BWT, are they really any different? Why?

## What is the main reason that BWT had an insignificant effect based on the SLR model, but a significant effect based on the MLR model?

## Compute sequential SS (type I SS)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
anova(lm(sbp~bwt+age))
anova(lm(sbp~bwt))
```

## Compute sequential SS (type I SS): order matters
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
anova(lm(sbp~age+bwt))
anova(lm(sbp~age))
```

## Numerically verify that, same as SAS, anova() uses F rather than F*
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(SeqSS = data.frame(anova(lm(sbp~bwt+age))))
SeqSS[,"Mean.Sq"]/SeqSS[,"F.value"]
```

## Compute partial SS (type III SS)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
library(car)
car::Anova(lm(sbp~bwt+age), type="III")
```


## Using the SS's, calculate how much would R$^2$ increase if AGE is added to a model that already contains BWT
\pause
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(seq_bwt_age = anova(lm(sbp~bwt+age)))
( SSY = sum( data.frame(seq_bwt_age)[,"Sum.Sq"] )  )
(diff_Rsquared=data.frame(seq_bwt_age)["age","Sum.Sq"]/SSY)
```


## Verify your answer using lm outputs
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~bwt+age))$r.squared
summary(lm(sbp~bwt))$r.squared
summary(lm(sbp~bwt+age))$r.squared - summary(lm(sbp~bwt))$r.squared
```

## Using the *extra* SS's, test whether BWT is associated with SBP, not accounting for AGE
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data.frame(seq_bwt_age)
```
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
## Compare to lm result
summary(lm(sbp~bwt))$coef["bwt",]
```
+ What is the reason for the difference?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~bwt))$sigma^2
```

## Are the t tests in lm(sbp~bwt+age) output sequential or partial? 
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~bwt+age))$coef
```
+ Verify your answer numerically\pause 
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(lm(sbp~bwt+age))$coef[,"t value"]^2
data.frame(car::Anova(lm(sbp~bwt+age), type="III"))[,"F.value"]
```