---
title: "Module_C"
author: Xu Shi
date: Sep 13, 2020
fontsize: 10pt
output: 
    beamer_presentation
latexengine: xelatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 60)
library(here)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
setwd(here("slides/")) 
```
## Example 2: 
+ A study is carried out in order to evaluate the relationship between birth weight (measured in g/100) and estriol levels (measured as mg/24h based on urine samples). Data were collected on a total of n=31 mothers-to-be and their newborns. A straight line regression model was used for analysis.
+ Read in the data file (estriol1.sas7bdat), which is stored as a permanent SAS set.
```{r, echo=T,eval=T}
library(sas7bdat)
data = read.sas7bdat(file="estriol1.sas7bdat")
attach(data)
```

## Obtain a listing of the variables. Check if there is any missing data.
+ Let's install and load a useful package called psych (or Hmisc)
```{r, warning=F, echo=T,eval=T}
library(psych)
```

## Obtain a listing of the variables. Check if there is any missing data.
+ The describe() function is handy
```{r, echo=T,eval=T,tidy.opts=list(blank=TRUE, width.cutoff=40),tidy=TRUE}
psych::describe(data)
```


## Fit an SLR model
```{r, mysize=T, tidy=T, size='\\footnotesize', warning=F, echo=T,eval=T}
m = lm(bwt~estriol); summary(m)
```
## By hand calculate parameter estiamtes 
```{r, echo=T,eval=T}
X=estriol; Y=bwt
Xbar = mean(X); Ybar = mean(Y)
n = length(Y)
( SSXY = sum((Y-Ybar)*(X-Xbar)) )
cov(X,Y)*(n-1)  # compare to SSXY
( SSX =  sum((X-Xbar)^2) )
var(X)*(n-1)  # compare to SSX
```

## By hand calculate parameter estiamtes (cont'd)
```{r, echo=T,eval=T}
(beta1hat = SSXY/SSX)              #beta1
as.numeric(coef(m)["estriol"])     #beta1 from lm
(beta0hat = Ybar-beta1hat*Xbar)    #beta0
as.numeric(coef(m)["(Intercept)"]) #beta0 from lm
```

## Obtain the F statistic and corresponding p value
+ Print ANOVA table
```{r, echo=T,eval=T}
anova(m)
```

## By hand calculate the F statistic and corresponding p value
```{r, echo=T,eval=T}
Yhat = m$fitted.values
( SSE = sum((Y-Yhat)^2) )
( SSR = sum((Yhat-Ybar)^2))
```
```{r  mysize=T, tidy=T, size='\\footnotesize', warning=F,echo=T,eval=T}
## Compare to ANOVA table
anova(m)
```


## By hand calculate the F statistic and corresponding p value (cont'd)
```{r, echo=T,eval=T}
MSE = SSE/(n-2) 
MSR = SSR/1
( Fstat = MSR/MSE )         ## F statistic
anova(m)$F[1]               ## F statistic from ANOVA table
1-pf(q=Fstat,df1=1,df2=n-2) ## P-value
anova(m)$Pr[1]              ## P-value from ANOVA table
```

## Estimate the marginal and conditional variance of birth weight.
+ Marginal variance of bwt
+ Variance of birth weight conditional on estriol level
```{r, echo=T,eval=T}
anova(m)  #var(bwt); sd(bwt)^2; sum((Y-Ybar)^2)/(n-1); MSE
```

## Does birth weight depend on estriol level? Carry out an appropriate t test
```{r, echo=T,eval=T}
sigma_squared = MSE                 
var_beta1hat = sigma_squared/SSX 
se_beta1hat = sqrt(var_beta1hat) 
t = beta1hat/se_beta1hat    ## t statistic
2*(1 - pt(q=t,df=n-2))      ## p-value
summary(m)$coef["estriol",] ## compare to lm result
```

## Use an F test to determine whether estriol levels are associated with birth weight
```{r, echo=T,eval=T}
( Fstat = MSR/MSE )         ## F statistic
1-pf(q=Fstat,df1=1,df2=n-2) ## P-value
```


## What percentage of the variation in birth weight is explained by estriol level?
```{r, echo=T,eval=T}
(R_squared = SSR / (SSR+SSE))## R-squared
summary(m)$r.squared         ## compare to R-squared from lm
```