---
title: "Module_L"
author: Xu Shi
date: Oct 24, 2020
fontsize: 10pt
output: 
    beamer_presentation
latexengine: pdflatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 70)
options(warn=-1)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
## Example: 
+ We return to the data set containing age (measured in days) and birth weight (BWT in ounces) as predictors of infant systolic blood pressure (SBP, mm Hg). 
+ Read in the data file (systolic1.csv). (n = 16)
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
rm(list=ls())
data = read.csv(file="systolic1.csv",header=T)
attach(data)
head(data)
```
## Fit a main effects model
+ sbp = $\beta_0$ + $\beta_A*$age + $\beta_B*$bwt + $\epsilon$
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m = lm(sbp~age+bwt)
```


## Which subject is most outlying with respect to the covariate space?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m.h = hatvalues(m)
Hmisc::describe(m.h)
m.h[which.max(m.h)]
```

## Using the moderate/large (0.2/0.5) criteria, how many subjects have moderate/large leverage?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m.h[which(m.h>0.5)]
m.h[which(m.h>0.2)]
```

## Using the $h_{ii}>2\overline{h}$ criterion, how many subjects are outlying with respect to $\boldsymbol{X}$?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m.h[which(m.h>2*mean(m.h))]
```


## (a) Output influence diagnostics.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m = lm(sbp~age+bwt)
m.dffits=dffits(m)
m.dfbeta=dfbeta(m)
m.D=cooks.distance(m)
m.covratio=covratio(m)
```
## (a) Plot Cook’s distance.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
n = nrow(data); p = m$rank
plot(m, which=4); abline(h=4/n,lty=2)
```


## (a) The plot() function is useful
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
par(mfrow=c(2,3)) #read more from ?plot.lm
plot(m,which=1);plot(m,which=2);plot(m,which=3)
plot(m,which=4);plot(m,which=5);plot(m,which=6)
```

## (a) Plot COVRATIO.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
which(   (abs(m.covratio-1) > 3*p/n)   )
plot(m.covratio); abline(1+3*p/n,0); abline(1-3*p/n,0)
```

## (a) Alternatively, the olsrr package
+ See more at https://olsrr.rsquaredacademy.com/
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_plot_dffits(m) # 2*sqrt(p/n) = 0.87
```

## (a) Alternatively, the olsrr package
+ See more at https://olsrr.rsquaredacademy.com/
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_plot_dfbetas(m) # 2/sqrt(n) = 0.5
```

## (a) Alternatively, the olsrr package
+ See more at https://olsrr.rsquaredacademy.com/
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
olsrr::ols_plot_cooksd_chart(m) # 4/n = 0.25
```

## (b) What subject has the greatest influence on $\widehat{\beta}_A$
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
head(m.dfbeta)
(ind.max.dfbeta = which.max(abs(m.dfbeta[,"age"])))
m.dfbeta[ind.max.dfbeta,]
```

## (c) For the subject identified in (b), approximate the value of $\widehat{\beta}_A$ if that subject was deleted
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
se.betaA = summary(m)$coef["age","Std. Error"]
betak_minus_betakremove = m.dfbeta[,"age"]*se.betaA
coef(m)["age"]-betak_minus_betakremove[ind.max.dfbeta]
```

## (d) Fit the model without the subject identified in (b). Why $\widehat{\beta}_A$ is unequal to that predicted in (c)?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m.remove = lm(sbp~age+bwt,data=data[-ind.max.dfbeta,])
coef(m.remove)["age"]
coef(m)["age"]-betak_minus_betakremove[ind.max.dfbeta]
coef(m)["age"]
```


## (e) Overall, what subject has the greatest impact on the parameter estimates (in the original model)?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
which.max(m.D)
```

## (f.0) How would you determine which subject has the most influence on the precision of estimation of $\widehat{\boldsymbol{\beta}}$?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
which.max(abs(m.covratio-1))
```

## (f) How would you determine which subject has the most influence on $\widehat{\sigma}^2$?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m.h = hatvalues(m)
m.sigmaratio = m.covratio*(1-m.h)
(ind.max.sigma = which.max(m.sigmaratio))
```



## (g) Approximate the value of $\widehat{\sigma}^2$ if the subject with the highest COVARTIO was deleted.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(m.sigmaratio[ind.max.sigma])^(1/p)*(summary(m)$sigma)^2
```

## (h) Re-fit the model, deleting the subject identified in (g). Compare your answer to that predicted in (g).
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m.remove2 = lm(sbp~age+bwt,data=data[-ind.max.sigma,])
(summary(m.remove2)$sigma)^2
(m.sigmaratio[ind.max.sigma])^(1/p)*(summary(m)$sigma)^2
(summary(m)$sigma)^2
```