---
title: "Module_E"
author: Xu Shi
date: Sep 24, 2021
fontsize: 10pt
output: 
    beamer_presentation
latexengine: xelatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 60)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
```
## Example: 
+ A study was carried out to examine the relationship between body weight (measured in pounds) and covariates age (years) and height (inches) among (n=12) children age 6-12.
+ Read in the data file (exampleE.csv).
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
data = read.csv(file="exampleE.csv",header=T)
attach(data)
```

## Using SLR, estimate the effect of age on weight
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m_age = lm(weight~age); summary(m_age)
```
## Using SLR, estimate the effect of age on weight
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
confint(m_age)
```

## Using SLR, estimate the effect of height on weight
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m_ht = lm(weight~height); summary(m_ht)
```
## Using SLR, estimate the effect of height on weight
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
confint(m_ht)
```

## Fit a multiple regression model with both height and age
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m_age_ht = lm(weight~age+height); summary(m_age_ht)
```
## Fit a multiple regression model with both height and age
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
confint(m_age_ht)
```

##  Compare the parameter estimates from SLR and MLR
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(m_age)$coef
summary(m_ht)$coef
summary(m_age_ht)$coef
```

##  Compare the parameter estimates from SLR and MLR
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
summary(m_age)$adj.r.squared
summary(m_ht)$adj.r.squared
summary(m_age_ht)$adj.r.squared
```


##  By-hand compute point estimates and make inference
```{r , echo=T,eval=T}
#### define design matrix, outcome, dimensions ####
X = model.matrix(weight~age+height)
Y = weight
n = nrow(X)
p = ncol(X)
#### Estimation: betahat and var(betahat) ####
betahat = solve(t(X)%*%X)%*%t(X)%*%Y 
Yhat = X%*%betahat                                    
epsilonhat = Y - Yhat ## residual
## estimated sigma^2
sigma_squared = t(epsilonhat)%*%epsilonhat/(n-p)      
## variance of betahat
var_betahat = diag( solve(t(X)%*%X) )*c(sigma_squared)
se_betahat = sqrt(var_betahat)  ## se of betahat      
#### Inference: t statistic and p val for H0: beta=0 ####
t_statistic = c(betahat/se_betahat) 
p_value = c(2*( 1-pt(q=abs(t_statistic),df=n-p) )) 
```

##  Compare results from by-hand calculation and lm()
+ By hand results
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(by_hand = cbind(Estimate=c(betahat),
                 Std_Err=se_betahat, 
                 t_statistic=t_statistic,
                 p_value=p_value))
```
+ Results from lm()
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
(from_lm = summary(m_age_ht)$coef)
```