---
title: "Module M"
author: Nicky
date: Nov 28, 2021
fontsize: 10pt
output: 
    beamer_presentation
latexengine: pdflatex
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 70)
options(warn=-1)
```
```{r  mysize=T, tidy=T, size='\\tiny', warning=F,echo=FALSE,eval=T}
knitr::knit_hooks$set(mysize = 
  function(before, options, envir){
    if (before) return(options$size)
  }
)
knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```
## Example: 
+ A data set is assembled from (n=17) hospitals in order to construct a model to predict manpower requirements (MANPOWER) as a function of the following factors: average number of patients admitted per day (PAT-DAYS), number of x- rays taken (X-RAYS), total number of days patients spent in bed (BED-DAYS), and average number of days stay per admission (AVG-STAY). One month’s worth of experience is recorded for each hospital.
+ Read in the data file (systolic1.csv).
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
rm(list=ls())
data = read.csv(file="hosp1.csv",header=T)
attach(data)
head(data)
```
## Compute the correlation matrix of the four covariates. 
+ Do you have any concerns? Which pair-wise correlation appears to be the most problematic?
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
cor(data[c("pat_days", "x_rays", "bed_days", "avg_stay")])
# psych::pairs.panels(data[,
#     c("pat_days", "x_rays", "bed_days", "avg_stay")])
```

## What are the limitations of your assessment via bivariate analysis?


## Fit a model containing each of the afore-listed factors. Compute the VIFs.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m = lm(manpower~pat_days+x_rays+bed_days+avg_stay)
car::vif(m)
```
+ Which covariate has the highest % of its variation explained by other covariates in the mode?


## Re-fit the model without BED-DAYS. Compare the R$^2$ to that of the original model.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m2 = lm(manpower~pat_days+x_rays+avg_stay)
c(summary(m)$r.squared, summary(m)$adj.r.squared)
c(summary(m2)$r.squared, summary(m2)$adj.r.squared)
```


## Re-fit the model without PAT-DAYS. Compare the R$^2$ to that of the original model.
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
m3 = lm(manpower~bed_days+x_rays+avg_stay)
c(summary(m)$r.squared, summary(m)$adj.r.squared)
c(summary(m2)$r.squared, summary(m2)$adj.r.squared)
c(summary(m3)$r.squared, summary(m3)$adj.r.squared)
```

## Based on your work thus far, what is a solution to the collinearity issue previously identified?
\pause
```{r mysize=T,size='\\footnotesize', echo=T,eval=T}
car::vif(m)
car::vif(m2)
car::vif(m3)
```
